---
title: "37-lncRNA-correlation"
output: html_document
date: "2025-12-03"
---

# lncRNA-Gene Expression Correlation Analysis

This analysis identifies which long non-coding RNAs (lncRNAs) have significant correlations 
(both positive and negative) with gene expression across three coral species:
- Acropora pulchra
- Porites evermanni  
- Pocillopora tuahiniensis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(corrr)
library(pheatmap)
library(ggplot2)
library(scales)
```

# Load Data

## Gene Expression Count Matrices

```{r load-gene-expression}
# Acropora pulchra
apul_gene <- read_csv(
  "https://gannet.fish.washington.edu/gitrepos/urol-e5/timeseries_molecular/D-Apul/output/02.20-D-Apul-RNAseq-alignment-HiSat2/apul-gene_count_matrix.csv",
  show_col_types = FALSE
)

# Porites evermanni
peve_gene <- read_csv(
  "https://gannet.fish.washington.edu/gitrepos/urol-e5/timeseries_molecular/E-Peve/output/02.20-E-Peve-RNAseq-alignment-HiSat2/peve-gene_count_matrix.csv",
  show_col_types = FALSE
)

# Pocillopora tuahiniensis
ptua_gene <- read_csv(
  "https://gannet.fish.washington.edu/gitrepos/urol-e5/timeseries_molecular/F-Ptua/output/02.20-F-Ptua-RNAseq-alignment-HiSat2/ptua-gene_count_matrix.csv",
  show_col_types = FALSE
)

cat("Gene expression matrices loaded:\n")
cat("Apul genes:", nrow(apul_gene), "x", ncol(apul_gene), "\n")
cat("Peve genes:", nrow(peve_gene), "x", ncol(peve_gene), "\n")
cat("Ptua genes:", nrow(ptua_gene), "x", ncol(ptua_gene), "\n")
```

## lncRNA Count Matrices

```{r load-lncrna}
# Acropora pulchra lncRNA
apul_lncrna <- read_tsv(
  "https://gannet.fish.washington.edu/v1_web/owlshell/bu-github/timeseries_molecular/D-Apul/output/31.5-Apul-lncRNA-discovery/lncRNA_counts.clean.filtered.txt",
  show_col_types = FALSE
)

# Porites evermanni lncRNA
peve_lncrna <- read_tsv(
  "https://gannet.fish.washington.edu/v1_web/owlshell/bu-github/timeseries_molecular/E-Peve/output/12-Peve-lncRNA-discovery/lncRNA_counts.clean.filtered.txt",
  show_col_types = FALSE
)

# Pocillopora tuahiniensis lncRNA
ptua_lncrna <- read_tsv(
  "https://raw.githubusercontent.com/urol-e5/timeseries_molecular/refs/heads/main/F-Ptua/output/06-Ptua-lncRNA-discovery/lncRNA_counts.clean.filtered.txt",
  show_col_types = FALSE
)

cat("lncRNA matrices loaded:\n")
cat("Apul lncRNAs:", nrow(apul_lncrna), "x", ncol(apul_lncrna), "\n")
cat("Peve lncRNAs:", nrow(peve_lncrna), "x", ncol(peve_lncrna), "\n")
cat("Ptua lncRNAs:", nrow(ptua_lncrna), "x", ncol(ptua_lncrna), "\n")
```

## Biomineralization Gene Lists

```{r load-biomin}
# Biomin gene lists for highlighting
apul_biomin <- read_csv(
  "https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/apul_biomin_counts.csv",
  show_col_types = FALSE
)

peve_biomin <- read_csv(
  "https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/peve_biomin_counts.csv",
  show_col_types = FALSE
)

ptua_biomin <- read_csv(
  "https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/ptua_biomin_counts.csv",
  show_col_types = FALSE
)

cat("Biomin genes loaded:\n")
cat("Apul biomin genes:", nrow(apul_biomin), "\n")
cat("Peve biomin genes:", nrow(peve_biomin), "\n")
cat("Ptua biomin genes:", nrow(ptua_biomin), "\n")
```

# Helper Functions

```{r helper-functions}
# Function to calculate correlations between lncRNA and gene expression
calculate_correlations <- function(gene_df, lncrna_df, gene_id_col = "gene_id", 
                                   lncrna_id_col = NULL, min_counts = 10) {
  
  # Identify ID columns
  if (is.null(lncrna_id_col)) {
    lncrna_id_col <- names(lncrna_df)[1]
  }
  
  # Get sample columns (shared between both matrices)
  gene_samples <- setdiff(names(gene_df), gene_id_col)
  lncrna_samples <- setdiff(names(lncrna_df), lncrna_id_col)
  shared_samples <- intersect(gene_samples, lncrna_samples)
  
  cat("Shared samples between gene and lncRNA data:", length(shared_samples), "\n")
  
  if (length(shared_samples) < 3) {
    warning("Not enough shared samples for correlation analysis")
    return(NULL)
  }
  
  # Filter to shared samples
  gene_mat <- gene_df %>%
    select(all_of(c(gene_id_col, shared_samples))) %>%
    column_to_rownames(gene_id_col) %>%
    as.matrix()
  
  lncrna_mat <- lncrna_df %>%
    select(all_of(c(lncrna_id_col, shared_samples))) %>%
    column_to_rownames(lncrna_id_col) %>%
    as.matrix()
  
  # Filter low-expression genes and lncRNAs
  gene_keep <- rowSums(gene_mat) >= min_counts
  lncrna_keep <- rowSums(lncrna_mat) >= min_counts
  
  gene_mat <- gene_mat[gene_keep, ]
  lncrna_mat <- lncrna_mat[lncrna_keep, ]
  
  cat("Genes after filtering:", nrow(gene_mat), "\n")
  cat("lncRNAs after filtering:", nrow(lncrna_mat), "\n")
  
  # Log2 transform (add 1 pseudocount)
  gene_log <- log2(gene_mat + 1)
  lncrna_log <- log2(lncrna_mat + 1)
  
  # Calculate correlations
  n_lncrna <- nrow(lncrna_log)
  n_genes <- nrow(gene_log)
  n_samples <- ncol(gene_log)
  
  cat("Calculating", n_lncrna * n_genes, "correlations...\n")
  
  # Pre-allocate results
  results <- vector("list", n_lncrna)
  
  for (i in seq_len(n_lncrna)) {
    lncrna_id <- rownames(lncrna_log)[i]
    lncrna_expr <- lncrna_log[i, ]
    
    cors <- apply(gene_log, 1, function(gene_expr) {
      if (sd(gene_expr) == 0 | sd(lncrna_expr) == 0) {
        return(c(cor = NA, pval = NA))
      }
      test <- cor.test(lncrna_expr, gene_expr, method = "pearson")
      c(cor = test$estimate, pval = test$p.value)
    })
    
    cors_df <- as.data.frame(t(cors))
    cors_df$gene_id <- rownames(cors_df)
    cors_df$lncrna_id <- lncrna_id
    results[[i]] <- cors_df
  }
  
  # Combine results
  all_cors <- bind_rows(results)
  
  # Adjust p-values for multiple testing
  all_cors <- all_cors %>%
    filter(!is.na(pval)) %>%
    mutate(padj = p.adjust(pval, method = "BH")) %>%
    arrange(padj)
  
  return(all_cors)
}

# Function to identify significant correlations
get_significant_cors <- function(cor_df, padj_threshold = 0.05, cor_threshold = 0.5) {
  cor_df %>%
    filter(padj < padj_threshold & abs(cor) >= cor_threshold) %>%
    mutate(direction = ifelse(cor > 0, "positive", "negative"))
}
```

# Acropora pulchra Analysis

```{r apul-correlation, cache=TRUE}
# Check column names
cat("Apul gene columns:", head(names(apul_gene)), "\n")
cat("Apul lncRNA columns:", head(names(apul_lncrna)), "\n")

# Calculate correlations
apul_cors <- calculate_correlations(
  apul_gene, 
  apul_lncrna,
  gene_id_col = names(apul_gene)[1],
  min_counts = 10
)

# Get significant correlations
apul_sig <- get_significant_cors(apul_cors, padj_threshold = 0.05, cor_threshold = 0.5)

cat("\nSignificant lncRNA-gene correlations in A. pulchra:\n")
cat("Total significant:", nrow(apul_sig), "\n")
cat("Positive correlations:", sum(apul_sig$direction == "positive"), "\n")
cat("Negative correlations:", sum(apul_sig$direction == "negative"), "\n")
```

```{r apul-biomin-highlight}
# Mark biomin genes
apul_sig <- apul_sig %>%
  mutate(is_biomin = gene_id %in% apul_biomin$gene_id)

cat("\nBiomin genes with significant lncRNA correlations:\n")
apul_sig %>%
  filter(is_biomin) %>%
  arrange(padj) %>%
  head(20)
```

```{r apul-plot, fig.width=10, fig.height=8}
# Volcano-style plot
ggplot(apul_cors %>% filter(!is.na(cor)), 
       aes(x = cor, y = -log10(pval))) +
  geom_point(alpha = 0.1, color = "gray70", size = 0.5) +
  geom_point(data = apul_sig %>% filter(!is_biomin),
             aes(color = direction), alpha = 0.5, size = 1) +
  geom_point(data = apul_sig %>% filter(is_biomin),
             color = "black", size = 2.5, shape = 21, fill = "gold", stroke = 0.8) +
  scale_color_manual(values = c("positive" = "#E74C3C", "negative" = "#3498DB")) +
  labs(title = "A. pulchra: lncRNA-Gene Expression Correlations",
       subtitle = "Gold = biomineralization genes",
       x = "Pearson Correlation",
       y = "-log10(p-value)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r apul-top-lncrnas}
# Top lncRNAs by number of significant correlations
apul_top_lncrnas <- apul_sig %>%
  group_by(lncrna_id) %>%
  summarise(
    n_positive = sum(direction == "positive"),
    n_negative = sum(direction == "negative"),
    n_total = n(),
    n_biomin = sum(is_biomin),
    mean_cor = mean(cor)
  ) %>%
  arrange(desc(n_total))

cat("\nTop 20 lncRNAs by number of correlated genes:\n")
head(apul_top_lncrnas, 20)
```

# Porites evermanni Analysis

```{r peve-correlation, cache=TRUE}
# Check column names
cat("Peve gene columns:", head(names(peve_gene)), "\n")
cat("Peve lncRNA columns:", head(names(peve_lncrna)), "\n")

# Calculate correlations
peve_cors <- calculate_correlations(
  peve_gene, 
  peve_lncrna,
  gene_id_col = names(peve_gene)[1],
  min_counts = 10
)

# Get significant correlations
peve_sig <- get_significant_cors(peve_cors, padj_threshold = 0.05, cor_threshold = 0.5)

cat("\nSignificant lncRNA-gene correlations in P. evermanni:\n")
cat("Total significant:", nrow(peve_sig), "\n")
cat("Positive correlations:", sum(peve_sig$direction == "positive"), "\n")
cat("Negative correlations:", sum(peve_sig$direction == "negative"), "\n")
```

```{r peve-biomin-highlight}
# Clean gene IDs for matching
peve_sig <- peve_sig %>%
  mutate(gene_id_clean = str_replace(gene_id, "^gene-", "")) %>%
  mutate(is_biomin = gene_id_clean %in% peve_biomin$gene_id | gene_id %in% peve_biomin$gene_id)

cat("\nBiomin genes with significant lncRNA correlations:\n")
peve_sig %>%
  filter(is_biomin) %>%
  arrange(padj) %>%
  head(20)
```

```{r peve-plot, fig.width=10, fig.height=8}
# Volcano-style plot
ggplot(peve_cors %>% filter(!is.na(cor)), 
       aes(x = cor, y = -log10(pval))) +
  geom_point(alpha = 0.1, color = "gray70", size = 0.5) +
  geom_point(data = peve_sig %>% filter(!is_biomin),
             aes(color = direction), alpha = 0.5, size = 1) +
  geom_point(data = peve_sig %>% filter(is_biomin),
             color = "black", size = 2.5, shape = 21, fill = "gold", stroke = 0.8) +
  scale_color_manual(values = c("positive" = "#E74C3C", "negative" = "#3498DB")) +
  labs(title = "P. evermanni: lncRNA-Gene Expression Correlations",
       subtitle = "Gold = biomineralization genes",
       x = "Pearson Correlation",
       y = "-log10(p-value)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r peve-top-lncrnas}
# Top lncRNAs by number of significant correlations
peve_top_lncrnas <- peve_sig %>%
  group_by(lncrna_id) %>%
  summarise(
    n_positive = sum(direction == "positive"),
    n_negative = sum(direction == "negative"),
    n_total = n(),
    n_biomin = sum(is_biomin),
    mean_cor = mean(cor)
  ) %>%
  arrange(desc(n_total))

cat("\nTop 20 lncRNAs by number of correlated genes:\n")
head(peve_top_lncrnas, 20)
```

# Pocillopora tuahiniensis Analysis

```{r ptua-correlation, cache=TRUE}
# Check column names
cat("Ptua gene columns:", head(names(ptua_gene)), "\n")
cat("Ptua lncRNA columns:", head(names(ptua_lncrna)), "\n")

# Calculate correlations
ptua_cors <- calculate_correlations(
  ptua_gene, 
  ptua_lncrna,
  gene_id_col = names(ptua_gene)[1],
  min_counts = 10
)

# Get significant correlations
ptua_sig <- get_significant_cors(ptua_cors, padj_threshold = 0.05, cor_threshold = 0.5)

cat("\nSignificant lncRNA-gene correlations in P. tuahiniensis:\n")
cat("Total significant:", nrow(ptua_sig), "\n")
cat("Positive correlations:", sum(ptua_sig$direction == "positive"), "\n")
cat("Negative correlations:", sum(ptua_sig$direction == "negative"), "\n")
```

```{r ptua-biomin-highlight}
# Clean gene IDs for matching
ptua_sig <- ptua_sig %>%
  mutate(gene_id_clean = str_replace(gene_id, "^gene-", "")) %>%
  mutate(is_biomin = gene_id_clean %in% ptua_biomin$gene_id | gene_id %in% ptua_biomin$gene_id)

cat("\nBiomin genes with significant lncRNA correlations:\n")
ptua_sig %>%
  filter(is_biomin) %>%
  arrange(padj) %>%
  head(20)
```

```{r ptua-plot, fig.width=10, fig.height=8}
# Volcano-style plot
ggplot(ptua_cors %>% filter(!is.na(cor)), 
       aes(x = cor, y = -log10(pval))) +
  geom_point(alpha = 0.1, color = "gray70", size = 0.5) +
  geom_point(data = ptua_sig %>% filter(!is_biomin),
             aes(color = direction), alpha = 0.5, size = 1) +
  geom_point(data = ptua_sig %>% filter(is_biomin),
             color = "black", size = 2.5, shape = 21, fill = "gold", stroke = 0.8) +
  scale_color_manual(values = c("positive" = "#E74C3C", "negative" = "#3498DB")) +
  labs(title = "P. tuahiniensis: lncRNA-Gene Expression Correlations",
       subtitle = "Gold = biomineralization genes",
       x = "Pearson Correlation",
       y = "-log10(p-value)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r ptua-top-lncrnas}
# Top lncRNAs by number of significant correlations
ptua_top_lncrnas <- ptua_sig %>%
  group_by(lncrna_id) %>%
  summarise(
    n_positive = sum(direction == "positive"),
    n_negative = sum(direction == "negative"),
    n_total = n(),
    n_biomin = sum(is_biomin),
    mean_cor = mean(cor)
  ) %>%
  arrange(desc(n_total))

cat("\nTop 20 lncRNAs by number of correlated genes:\n")
head(ptua_top_lncrnas, 20)
```

# Summary Across Species

```{r summary-comparison, fig.width=12, fig.height=5}
# Create summary dataframe
summary_df <- data.frame(
  Species = c("A. pulchra", "P. evermanni", "P. tuahiniensis"),
  Positive = c(
    sum(apul_sig$direction == "positive"),
    sum(peve_sig$direction == "positive"),
    sum(ptua_sig$direction == "positive")
  ),
  Negative = c(
    sum(apul_sig$direction == "negative"),
    sum(peve_sig$direction == "negative"),
    sum(ptua_sig$direction == "negative")
  ),
  Biomin_associated = c(
    sum(apul_sig$is_biomin),
    sum(peve_sig$is_biomin),
    sum(ptua_sig$is_biomin)
  )
)

summary_df$Total <- summary_df$Positive + summary_df$Negative

print(summary_df)

# Plot comparison
summary_long <- summary_df %>%
  pivot_longer(cols = c(Positive, Negative), 
               names_to = "Direction", 
               values_to = "Count")

ggplot(summary_long, aes(x = Species, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Positive" = "#E74C3C", "Negative" = "#3498DB")) +
  labs(title = "Significant lncRNA-Gene Correlations Across Species",
       y = "Number of Significant Correlations") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Biomin Gene Focus

```{r biomin-focus, fig.width=14, fig.height=6}
# Combine biomin results across species
biomin_combined <- bind_rows(
  apul_sig %>% filter(is_biomin) %>% mutate(species = "A. pulchra"),
  peve_sig %>% filter(is_biomin) %>% mutate(species = "P. evermanni"),
  ptua_sig %>% filter(is_biomin) %>% mutate(species = "P. tuahiniensis")
)

if (nrow(biomin_combined) > 0) {
  # Plot biomin correlations
  ggplot(biomin_combined, aes(x = cor, y = reorder(gene_id, cor), fill = direction)) +
    geom_col() +
    facet_wrap(~species, scales = "free_y") +
    scale_fill_manual(values = c("positive" = "#E74C3C", "negative" = "#3498DB")) +
    labs(title = "Biomineralization Genes with Significant lncRNA Correlations",
         x = "Correlation Coefficient",
         y = "Gene ID") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 6))
} else {
  cat("No significant biomin-lncRNA correlations found\n")
}
```

# Save Results

```{r save-results}
# Create output directory
output_dir <- "../output/37-lncRNA-correlation"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# Save significant correlations for each species
write_csv(apul_sig, file.path(output_dir, "apul_lncrna_gene_correlations.csv"))
write_csv(peve_sig, file.path(output_dir, "peve_lncrna_gene_correlations.csv"))
write_csv(ptua_sig, file.path(output_dir, "ptua_lncrna_gene_correlations.csv"))

# Save top lncRNAs
write_csv(apul_top_lncrnas, file.path(output_dir, "apul_top_lncrnas.csv"))
write_csv(peve_top_lncrnas, file.path(output_dir, "peve_top_lncrnas.csv"))
write_csv(ptua_top_lncrnas, file.path(output_dir, "ptua_top_lncrnas.csv"))

# Save summary
write_csv(summary_df, file.path(output_dir, "species_summary.csv"))

cat("Results saved to:", output_dir, "\n")
```

# Session Info

```{r session-info}
sessionInfo()
```

