---
title: "49-AS: Alternative Splicing Analysis"
author: "Timeseries Molecular Calcification Project"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)

# Required packages
library(ballgown)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(here)
```

# Overview

This notebook performs alternative splicing (AS) analysis for three coral species:

- **Acropora pulchra** (Apul)
- **Porites evermanni** (Peve)
- **Pocillopora tuahiniensis** (Ptua)

## Background

Each sample's StringTie/Ballgown output directory contains:

| File | Description |
|------|-------------|
| `e_data.ctab` | Exon-level counts and FPKM |
| `i_data.ctab` | Intron/junction coverage |
| `t_data.ctab` | Transcript-level FPKM |
| `e2t.ctab` | Exon-to-transcript mapping |
| `i2t.ctab` | Intron-to-transcript mapping |

Alternative splicing events are identified when:

1. Transcripts of the same gene change in relative abundance across conditions (timepoints)
2. Certain exons or splice junctions show condition-specific expression patterns

## Analysis Strategies

We employ three complementary strategies:

1. **Strategy 1**: Differential Transcript Usage (DTU) — Identify genes whose isoform proportions change across timepoints
2. **Strategy 2**: Differential Exon Usage — Identify exons with expression changes independent of whole-gene expression
3. **Strategy 3**: Junction-Level Tests — Identify differential splice junction usage

```{r paths}
# Base directory
base_dir <- here::here("M-multi-species")

# Species-specific RNA-seq alignment output directories
apul_rnaseq_dir <- here::here("D-Apul/output/02.20-D-Apul-RNAseq-alignment-HiSat2")
peve_rnaseq_dir <- here::here("E-Peve/output/02.20-E-Peve-RNAseq-alignment-HiSat2")
ptua_rnaseq_dir <- here::here("F-Ptua/output/02.20-F-Ptua-RNAseq-alignment-HiSat2")

# Master metadata
metadata_file <- here::here("master-molecular-metadata.csv")

# Output directory for AS results
output_dir <- here::here("M-multi-species/output/49-AS")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Read master metadata
master_metadata <- read_csv(metadata_file, show_col_types = FALSE)
```

# Helper Functions

## Create Sample Metadata

```{r helper-create-metadata}
#' Create sample metadata for ballgown from master metadata
#' @param rnaseq_dir Path to RNA-seq alignment output directory
#' @param species_pattern Regex pattern to identify species (e.g., "ACR", "POR", "POC")
#' @param master_meta Master metadata dataframe
#' @return Dataframe with sample_id, path, colony, timepoint, site
create_sample_metadata <- function(rnaseq_dir, species_pattern, master_meta) {
  # Get sample directories
  all_subdirs <- list.dirs(rnaseq_dir, full.names = FALSE, recursive = FALSE)
  sample_ids <- all_subdirs[grepl(species_pattern, all_subdirs)]
  
  # Filter for only sample directories (those containing ctab files)
  valid_samples <- sample_ids[sapply(sample_ids, function(s) {
    file.exists(file.path(rnaseq_dir, s, "e_data.ctab"))
  })]
  
  # Create metadata dataframe
  sample_meta <- data.frame(
    sample_id = valid_samples,
    path = file.path(rnaseq_dir, valid_samples),
    stringsAsFactors = FALSE
  )
  
  # Parse colony and timepoint from sample_id (e.g., "ACR-139-TP1" -> colony="ACR-139", timepoint="TP1")
  sample_meta <- sample_meta %>%
    mutate(
      colony = sub("-TP[1-4]$", "", sample_id),
      timepoint = sub(".*-(TP[1-4])$", "\\1", sample_id)
    )
  
  # Join with master metadata for site information
  sample_meta <- sample_meta %>%
    left_join(
      master_meta %>% 
        select(colony, timepoint, site, species) %>%
        distinct(),
      by = c("colony", "timepoint")
    )
  
  # Set rownames for ballgown pData
  rownames(sample_meta) <- sample_meta$sample_id
  
  sample_meta
}
```

## Load Ballgown Object

```{r helper-load-ballgown}
#' Load ballgown object from sample directories
#' @param sample_meta Sample metadata with 'path' column
#' @return ballgown object
load_ballgown_object <- function(sample_meta) {
  # ballgown requires directories containing ctab files
  bg <- ballgown(
    samples = sample_meta$path,
    meas = "all",
    pData = sample_meta
  )
  return(bg)
}
```

# Acropora pulchra (Apul) Analysis

## Load Data

```{r apul-load}
# Create sample metadata for Apul
apul_sample_meta <- create_sample_metadata(
  rnaseq_dir = apul_rnaseq_dir,
  species_pattern = "^ACR-",
  master_meta = master_metadata
)

cat("Apul samples loaded:", nrow(apul_sample_meta), "\n")
cat("Timepoint distribution:\n")
print(table(apul_sample_meta$timepoint))

# Load ballgown object
apul_bg <- load_ballgown_object(apul_sample_meta)
```

## Strategy 1: Differential Transcript Usage (DTU)

Identify genes whose isoform proportions change across timepoints.

```{r apul-dtu}
# Perform differential expression on transcripts across timepoints
apul_transcript_results <- stattest(
  apul_bg,
  feature = "transcript",
  meas = "FPKM",
  covariate = "timepoint",
  getFC = TRUE
)

# Add gene information to transcript results
# Get transcript data from ballgown object
apul_t_data <- texpr(apul_bg, meas = "all")
apul_transcript_results <- apul_transcript_results %>%
  mutate(id = as.integer(id)) %>%
  left_join(
    data.frame(
      id = as.integer(apul_t_data$t_id),
      gene_id = apul_t_data$gene_id,
      t_name = apul_t_data$t_name
    ),
    by = "id"
  )

# Identify genes with multiple differentially expressed transcripts
apul_dtu_candidates <- apul_transcript_results %>%
  filter(!is.na(qval)) %>%
  group_by(gene_id) %>%
  summarize(
    n_transcripts = n(),
    n_sig = sum(qval < 0.05, na.rm = TRUE),
    min_qval = min(qval, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_transcripts >= 2, n_sig >= 1) %>%
  arrange(min_qval)

cat("Apul DTU candidate genes:", nrow(apul_dtu_candidates), "\n")
head(apul_dtu_candidates, 20)
```

## Strategy 2: Differential Exon Usage

Identify exons whose expression changes independently of whole-gene expression.

```{r apul-exon}
# Perform differential expression on exons
apul_exon_results <- stattest(
  apul_bg,
  feature = "exon",
  meas = "rcount",
  covariate = "timepoint"
)

# Filter significant exons
apul_exon_sig <- apul_exon_results %>%
  filter(qval < 0.05) %>%
  arrange(qval)

cat("Apul significant exons:", nrow(apul_exon_sig), "\n")
head(apul_exon_sig, 20)
```

## Strategy 3: Junction-Level Tests

Identify differential splice junction usage.

```{r apul-junction}
# Perform differential expression on introns/junctions
apul_junction_results <- stattest(
  apul_bg,
  feature = "intron",
  meas = "rcount",
  covariate = "timepoint"
)

# Filter significant junctions
apul_junction_sig <- apul_junction_results %>%
  filter(qval < 0.05) %>%
  arrange(qval)

cat("Apul significant junctions:", nrow(apul_junction_sig), "\n")
head(apul_junction_sig, 20)
```

## Integrate Apul Results

```{r apul-integrate}
# Get gene IDs from exon results by mapping e_id to gene_id via transcript data
apul_e2t <- eexpr(apul_bg, meas = "all")
apul_exon_gene_map <- data.frame(
  e_id = seq_len(nrow(apul_e2t))
)
# For exons, we need to map through the structure - use indexes from significant results
if (nrow(apul_exon_sig) > 0) {
  apul_exon_gene_ids <- unique(apul_t_data$gene_id[apul_exon_sig$id])
} else {
  apul_exon_gene_ids <- character(0)
}

# Get gene IDs from junction results
if (nrow(apul_junction_sig) > 0) {
  apul_junction_gene_ids <- unique(apul_t_data$gene_id[apul_junction_sig$id])
} else {
  apul_junction_gene_ids <- character(0)
}

# Combine all AS candidate genes
apul_AS_genes <- unique(c(
  apul_dtu_candidates$gene_id,
  apul_exon_gene_ids,
  apul_junction_gene_ids
))
apul_AS_genes <- apul_AS_genes[!is.na(apul_AS_genes)]

cat("Apul total AS candidate genes:", length(apul_AS_genes), "\n")

# Create summary dataframe
apul_AS_summary <- data.frame(
  gene_id = apul_AS_genes,
  in_dtu = apul_AS_genes %in% apul_dtu_candidates$gene_id,
  in_exon = apul_AS_genes %in% apul_exon_gene_ids,
  in_junction = apul_AS_genes %in% apul_junction_gene_ids
) %>%
  mutate(n_evidence = in_dtu + in_exon + in_junction) %>%
  arrange(desc(n_evidence))

cat("\nEvidence distribution:\n")
print(table(apul_AS_summary$n_evidence))
```

# Porites evermanni (Peve) Analysis

## Load Data

```{r peve-load}
# Create sample metadata for Peve
peve_sample_meta <- create_sample_metadata(
  rnaseq_dir = peve_rnaseq_dir,
  species_pattern = "^POR-",
  master_meta = master_metadata
)

cat("Peve samples loaded:", nrow(peve_sample_meta), "\n")
cat("Timepoint distribution:\n")
print(table(peve_sample_meta$timepoint))

# Load ballgown object
peve_bg <- load_ballgown_object(peve_sample_meta)
```

## Strategy 1: Differential Transcript Usage (DTU)

```{r peve-dtu}
# Perform differential expression on transcripts across timepoints
peve_transcript_results <- stattest(
  peve_bg,
  feature = "transcript",
  meas = "FPKM",
  covariate = "timepoint",
  getFC = TRUE
)

# Add gene information to transcript results
peve_t_data <- texpr(peve_bg, meas = "all")
peve_transcript_results <- peve_transcript_results %>%
  mutate(id = as.integer(id)) %>%
  left_join(
    data.frame(
      id = as.integer(peve_t_data$t_id),
      gene_id = peve_t_data$gene_id,
      t_name = peve_t_data$t_name
    ),
    by = "id"
  )

# Identify genes with multiple differentially expressed transcripts
peve_dtu_candidates <- peve_transcript_results %>%
  filter(!is.na(qval)) %>%
  group_by(gene_id) %>%
  summarize(
    n_transcripts = n(),
    n_sig = sum(qval < 0.05, na.rm = TRUE),
    min_qval = min(qval, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_transcripts >= 2, n_sig >= 1) %>%
  arrange(min_qval)

cat("Peve DTU candidate genes:", nrow(peve_dtu_candidates), "\n")
head(peve_dtu_candidates, 20)
```

## Strategy 2: Differential Exon Usage

```{r peve-exon}
# Perform differential expression on exons
peve_exon_results <- stattest(
  peve_bg,
  feature = "exon",
  meas = "rcount",
  covariate = "timepoint"
)

# Filter significant exons
peve_exon_sig <- peve_exon_results %>%
  filter(qval < 0.05) %>%
  arrange(qval)

cat("Peve significant exons:", nrow(peve_exon_sig), "\n")
head(peve_exon_sig, 20)
```

## Strategy 3: Junction-Level Tests

```{r peve-junction}
# Perform differential expression on introns/junctions
peve_junction_results <- stattest(
  peve_bg,
  feature = "intron",
  meas = "rcount",
  covariate = "timepoint"
)

# Filter significant junctions
peve_junction_sig <- peve_junction_results %>%
  filter(qval < 0.05) %>%
  arrange(qval)

cat("Peve significant junctions:", nrow(peve_junction_sig), "\n")
head(peve_junction_sig, 20)
```

## Integrate Peve Results

```{r peve-integrate}
# Get gene IDs from exon results
if (nrow(peve_exon_sig) > 0) {
  peve_exon_gene_ids <- unique(peve_t_data$gene_id[peve_exon_sig$id])
} else {
  peve_exon_gene_ids <- character(0)
}

# Get gene IDs from junction results
if (nrow(peve_junction_sig) > 0) {
  peve_junction_gene_ids <- unique(peve_t_data$gene_id[peve_junction_sig$id])
} else {
  peve_junction_gene_ids <- character(0)
}

# Combine all AS candidate genes
peve_AS_genes <- unique(c(
  peve_dtu_candidates$gene_id,
  peve_exon_gene_ids,
  peve_junction_gene_ids
))
peve_AS_genes <- peve_AS_genes[!is.na(peve_AS_genes)]

cat("Peve total AS candidate genes:", length(peve_AS_genes), "\n")

# Create summary dataframe
peve_AS_summary <- data.frame(
  gene_id = peve_AS_genes,
  in_dtu = peve_AS_genes %in% peve_dtu_candidates$gene_id,
  in_exon = peve_AS_genes %in% peve_exon_gene_ids,
  in_junction = peve_AS_genes %in% peve_junction_gene_ids
) %>%
  mutate(n_evidence = in_dtu + in_exon + in_junction) %>%
  arrange(desc(n_evidence))

cat("\nEvidence distribution:\n")
print(table(peve_AS_summary$n_evidence))
```

# Pocillopora tuahiniensis (Ptua) Analysis

## Load Data

```{r ptua-load}
# Create sample metadata for Ptua
ptua_sample_meta <- create_sample_metadata(
  rnaseq_dir = ptua_rnaseq_dir,
  species_pattern = "^POC-",
  master_meta = master_metadata
)

cat("Ptua samples loaded:", nrow(ptua_sample_meta), "\n")
cat("Timepoint distribution:\n")
print(table(ptua_sample_meta$timepoint))

# Load ballgown object
ptua_bg <- load_ballgown_object(ptua_sample_meta)
```

## Strategy 1: Differential Transcript Usage (DTU)

```{r ptua-dtu}
# Perform differential expression on transcripts across timepoints
ptua_transcript_results <- stattest(
  ptua_bg,
  feature = "transcript",
  meas = "FPKM",
  covariate = "timepoint",
  getFC = TRUE
)

# Add gene information to transcript results
ptua_t_data <- texpr(ptua_bg, meas = "all")
ptua_transcript_results <- ptua_transcript_results %>%
  mutate(id = as.integer(id)) %>%
  left_join(
    data.frame(
      id = as.integer(ptua_t_data$t_id),
      gene_id = ptua_t_data$gene_id,
      t_name = ptua_t_data$t_name
    ),
    by = "id"
  )

# Identify genes with multiple differentially expressed transcripts
ptua_dtu_candidates <- ptua_transcript_results %>%
  filter(!is.na(qval)) %>%
  group_by(gene_id) %>%
  summarize(
    n_transcripts = n(),
    n_sig = sum(qval < 0.05, na.rm = TRUE),
    min_qval = min(qval, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_transcripts >= 2, n_sig >= 1) %>%
  arrange(min_qval)

cat("Ptua DTU candidate genes:", nrow(ptua_dtu_candidates), "\n")
head(ptua_dtu_candidates, 20)
```

## Strategy 2: Differential Exon Usage

```{r ptua-exon}
# Perform differential expression on exons
ptua_exon_results <- stattest(
  ptua_bg,
  feature = "exon",
  meas = "rcount",
  covariate = "timepoint"
)

# Filter significant exons
ptua_exon_sig <- ptua_exon_results %>%
  filter(qval < 0.05) %>%
  arrange(qval)

cat("Ptua significant exons:", nrow(ptua_exon_sig), "\n")
head(ptua_exon_sig, 20)
```

## Strategy 3: Junction-Level Tests

```{r ptua-junction}
# Perform differential expression on introns/junctions
ptua_junction_results <- stattest(
  ptua_bg,
  feature = "intron",
  meas = "rcount",
  covariate = "timepoint"
)

# Filter significant junctions
ptua_junction_sig <- ptua_junction_results %>%
  filter(qval < 0.05) %>%
  arrange(qval)

cat("Ptua significant junctions:", nrow(ptua_junction_sig), "\n")
head(ptua_junction_sig, 20)
```

## Integrate Ptua Results

```{r ptua-integrate}
# Get gene IDs from exon results
if (nrow(ptua_exon_sig) > 0) {
  ptua_exon_gene_ids <- unique(ptua_t_data$gene_id[ptua_exon_sig$id])
} else {
  ptua_exon_gene_ids <- character(0)
}

# Get gene IDs from junction results
if (nrow(ptua_junction_sig) > 0) {
  ptua_junction_gene_ids <- unique(ptua_t_data$gene_id[ptua_junction_sig$id])
} else {
  ptua_junction_gene_ids <- character(0)
}

# Combine all AS candidate genes
ptua_AS_genes <- unique(c(
  ptua_dtu_candidates$gene_id,
  ptua_exon_gene_ids,
  ptua_junction_gene_ids
))
ptua_AS_genes <- ptua_AS_genes[!is.na(ptua_AS_genes)]

cat("Ptua total AS candidate genes:", length(ptua_AS_genes), "\n")

# Create summary dataframe
ptua_AS_summary <- data.frame(
  gene_id = ptua_AS_genes,
  in_dtu = ptua_AS_genes %in% ptua_dtu_candidates$gene_id,
  in_exon = ptua_AS_genes %in% ptua_exon_gene_ids,
  in_junction = ptua_AS_genes %in% ptua_junction_gene_ids
) %>%
  mutate(n_evidence = in_dtu + in_exon + in_junction) %>%
  arrange(desc(n_evidence))

cat("\nEvidence distribution:\n")
print(table(ptua_AS_summary$n_evidence))
```

# Cross-Species Comparison

## Summary Statistics

```{r cross-species-summary}
# Create summary table
as_summary <- data.frame(
  Species = c("Apul", "Peve", "Ptua"),
  Total_AS_Genes = c(length(apul_AS_genes), length(peve_AS_genes), length(ptua_AS_genes)),
  DTU_Candidates = c(nrow(apul_dtu_candidates), nrow(peve_dtu_candidates), nrow(ptua_dtu_candidates)),
  Sig_Exons = c(nrow(apul_exon_sig), nrow(peve_exon_sig), nrow(ptua_exon_sig)),
  Sig_Junctions = c(nrow(apul_junction_sig), nrow(peve_junction_sig), nrow(ptua_junction_sig))
)

print(as_summary)
```

# Data Quality Diagnostics

This section investigates potential causes for dramatic differences in AS detection across species.

## Exon Expression Distribution

```{r diag-exon-expression, fig.width=12, fig.height=8}
# Extract exon-level coverage for each species (FPKM not available for exons)
apul_exon_cov <- eexpr(apul_bg, meas = "cov")
peve_exon_cov <- eexpr(peve_bg, meas = "cov")
ptua_exon_cov <- eexpr(ptua_bg, meas = "cov")

# Summary statistics
cat("=== Exon Coverage Summary ===\n\n")

cat("Apul:\n")
cat("  Total exons:", nrow(apul_exon_cov), "\n")
cat("  Mean coverage:", round(mean(as.matrix(apul_exon_cov), na.rm = TRUE), 3), "\n")
cat("  Median coverage:", round(median(as.matrix(apul_exon_cov), na.rm = TRUE), 3), "\n")
cat("  % zero:", round(100 * sum(as.matrix(apul_exon_cov) == 0, na.rm = TRUE) / length(as.matrix(apul_exon_cov)), 2), "\n\n")

cat("Peve:\n")
cat("  Total exons:", nrow(peve_exon_cov), "\n")
cat("  Mean coverage:", round(mean(as.matrix(peve_exon_cov), na.rm = TRUE), 3), "\n")
cat("  Median coverage:", round(median(as.matrix(peve_exon_cov), na.rm = TRUE), 3), "\n")
cat("  % zero:", round(100 * sum(as.matrix(peve_exon_cov) == 0, na.rm = TRUE) / length(as.matrix(peve_exon_cov)), 2), "\n\n")

cat("Ptua:\n")
cat("  Total exons:", nrow(ptua_exon_cov), "\n")
cat("  Mean coverage:", round(mean(as.matrix(ptua_exon_cov), na.rm = TRUE), 3), "\n")
cat("  Median coverage:", round(median(as.matrix(ptua_exon_cov), na.rm = TRUE), 3), "\n")
cat("  % zero:", round(100 * sum(as.matrix(ptua_exon_cov) == 0, na.rm = TRUE) / length(as.matrix(ptua_exon_cov)), 2), "\n")

# Create combined density plot
apul_cov_long <- data.frame(
  species = "Apul",
  log_cov = log10(as.vector(as.matrix(apul_exon_cov)) + 1)
)
peve_cov_long <- data.frame(
  species = "Peve", 
  log_cov = log10(as.vector(as.matrix(peve_exon_cov)) + 1)
)
ptua_cov_long <- data.frame(
  species = "Ptua",
  log_cov = log10(as.vector(as.matrix(ptua_exon_cov)) + 1)
)

cov_combined <- bind_rows(apul_cov_long, peve_cov_long, ptua_cov_long)

ggplot(cov_combined, aes(x = log_cov, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Distribution of Exon Coverage Across Species",
    subtitle = "log10(coverage + 1)",
    x = "log10(coverage + 1)",
    y = "Density"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

## P-value Distribution Diagnostics

```{r diag-pvalue-dist, fig.width=14, fig.height=5}
# P-value histograms - uniform distribution indicates no signal or well-calibrated null
par(mfrow = c(1, 3))

# Apul
hist(apul_exon_results$pval, breaks = 50, main = "Apul Exon P-values",
     xlab = "P-value", col = "steelblue", border = "white")
abline(h = nrow(apul_exon_results[!is.na(apul_exon_results$pval),]) / 50, 
       col = "red", lty = 2, lwd = 2)

# Peve  
hist(peve_exon_results$pval, breaks = 50, main = "Peve Exon P-values",
     xlab = "P-value", col = "coral", border = "white")
abline(h = nrow(peve_exon_results[!is.na(peve_exon_results$pval),]) / 50,
       col = "red", lty = 2, lwd = 2)

# Ptua
hist(ptua_exon_results$pval, breaks = 50, main = "Ptua Exon P-values",
     xlab = "P-value", col = "seagreen", border = "white")
abline(h = nrow(ptua_exon_results[!is.na(ptua_exon_results$pval),]) / 50,
       col = "red", lty = 2, lwd = 2)

cat("\n=== P-value Distribution Summary ===\n")
cat("Red dashed line = expected count under null hypothesis (uniform distribution)\n")
cat("Enrichment near 0 indicates true signal; uniform = no signal or data issue\n\n")

cat("Apul - Exons with p < 0.01:", sum(apul_exon_results$pval < 0.01, na.rm = TRUE), 
    "(", round(100 * sum(apul_exon_results$pval < 0.01, na.rm = TRUE) / sum(!is.na(apul_exon_results$pval)), 2), "%)\n")
cat("Peve - Exons with p < 0.01:", sum(peve_exon_results$pval < 0.01, na.rm = TRUE),
    "(", round(100 * sum(peve_exon_results$pval < 0.01, na.rm = TRUE) / sum(!is.na(peve_exon_results$pval)), 2), "%)\n")
cat("Ptua - Exons with p < 0.01:", sum(ptua_exon_results$pval < 0.01, na.rm = TRUE),
    "(", round(100 * sum(ptua_exon_results$pval < 0.01, na.rm = TRUE) / sum(!is.na(ptua_exon_results$pval)), 2), "%)\n")
```

## Exon Variance Comparison

```{r diag-variance, fig.width=12, fig.height=5}
# Calculate per-exon variance (coefficient of variation)
calc_cv <- function(cov_matrix) {
  apply(cov_matrix, 1, function(x) {
    m <- mean(x, na.rm = TRUE)
    s <- sd(x, na.rm = TRUE)
    if (m > 0) s / m else NA
  })
}

apul_cv <- calc_cv(apul_exon_cov)
peve_cv <- calc_cv(peve_exon_cov)
ptua_cv <- calc_cv(ptua_exon_cov)

cat("=== Coefficient of Variation (CV) Summary ===\n")
cat("Higher CV indicates more variable expression across samples\n\n")

cat("Apul - Median CV:", round(median(apul_cv, na.rm = TRUE), 3), 
    "Mean CV:", round(mean(apul_cv, na.rm = TRUE), 3), "\n")
cat("Peve - Median CV:", round(median(peve_cv, na.rm = TRUE), 3),
    "Mean CV:", round(mean(peve_cv, na.rm = TRUE), 3), "\n")
cat("Ptua - Median CV:", round(median(ptua_cv, na.rm = TRUE), 3),
    "Mean CV:", round(mean(ptua_cv, na.rm = TRUE), 3), "\n")

# CV distribution plot
cv_combined <- data.frame(
  species = c(rep("Apul", length(apul_cv)), 
              rep("Peve", length(peve_cv)), 
              rep("Ptua", length(ptua_cv))),
  cv = c(apul_cv, peve_cv, ptua_cv)
) %>%
  filter(!is.na(cv) & cv < 5)  # Filter extreme values for visualization

ggplot(cv_combined, aes(x = cv, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Distribution of Exon Expression Variability",
    subtitle = "Coefficient of Variation (CV) across samples - lower CV may explain fewer significant results",
    x = "Coefficient of Variation",
    y = "Density"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50")
```

## Sample-level Expression PCA

```{r diag-pca, fig.width=14, fig.height=5}
# PCA to visualize sample separation by timepoint
run_pca_plot <- function(bg_obj, sample_meta, species_name) {
  # Get transcript FPKM
  fpkm <- texpr(bg_obj, meas = "FPKM")
  
  # Filter low expression transcripts
  keep <- rowSums(fpkm > 1) >= 3
  fpkm_filt <- fpkm[keep, ]
  
  if (nrow(fpkm_filt) < 10) {
    cat(species_name, ": Not enough transcripts for PCA\n")
    return(NULL)
  }
  
  # Log transform
  log_fpkm <- log2(fpkm_filt + 1)
  
  # PCA
  pca_result <- prcomp(t(log_fpkm), center = TRUE, scale. = TRUE)
  
  # Extract PC scores
  pc_scores <- as.data.frame(pca_result$x[, 1:2])
  pc_scores$sample_id <- rownames(pc_scores)
  pc_scores$timepoint <- sample_meta$timepoint[match(pc_scores$sample_id, sample_meta$sample_id)]
  
  # Variance explained
  var_explained <- round(100 * summary(pca_result)$importance[2, 1:2], 1)
  
  ggplot(pc_scores, aes(x = PC1, y = PC2, color = timepoint)) +
    geom_point(size = 3) +
    labs(
      title = paste(species_name, "- Sample PCA"),
      subtitle = "Colored by timepoint - separation indicates temporal signal",
      x = paste0("PC1 (", var_explained[1], "%)"),
      y = paste0("PC2 (", var_explained[2], "%)")
    ) +
    theme_minimal() +
    scale_color_brewer(palette = "Set1")
}

# Create PCA plots
p1 <- run_pca_plot(apul_bg, apul_sample_meta, "Apul")
p2 <- run_pca_plot(peve_bg, peve_sample_meta, "Peve")
p3 <- run_pca_plot(ptua_bg, ptua_sample_meta, "Ptua")

# Combine plots if all exist
if (!is.null(p1) && !is.null(p2) && !is.null(p3)) {
  library(patchwork)
  p1 + p2 + p3
}
```

## NA Rate Comparison

```{r diag-na-rate}
cat("=== NA Rate in Statistical Tests ===\n")
cat("High NA rate may indicate low/zero counts or computational issues\n\n")

calc_na_rate <- function(results_df) {
  n_total <- nrow(results_df)
  n_na <- sum(is.na(results_df$pval))
  list(total = n_total, na_count = n_na, pct = round(100 * n_na / n_total, 2))
}

apul_na <- calc_na_rate(apul_exon_results)
peve_na <- calc_na_rate(peve_exon_results)
ptua_na <- calc_na_rate(ptua_exon_results)

na_summary <- data.frame(
  Species = c("Apul", "Peve", "Ptua"),
  Total_Exons = c(apul_na$total, peve_na$total, ptua_na$total),
  NA_Count = c(apul_na$na_count, peve_na$na_count, ptua_na$na_count),
  NA_Percent = c(apul_na$pct, peve_na$pct, ptua_na$pct)
)

print(na_summary)
```

## Timepoint Sample Balance

```{r diag-sample-balance}
cat("=== Sample Distribution by Timepoint ===\n")
cat("Imbalanced design may reduce statistical power\n\n")

cat("Apul:\n")
print(table(apul_sample_meta$timepoint))

cat("\nPeve:\n")
print(table(peve_sample_meta$timepoint))

cat("\nPtua:\n")
print(table(ptua_sample_meta$timepoint))
```

## Diagnostic Summary

```{r diag-summary}
cat("=== DIAGNOSTIC SUMMARY ===\n\n")

cat("1. EXPRESSION LEVELS:\n")
cat("   Compare mean/median FPKM and % zeros between species.\n")
cat("   Lower expression = less power to detect differences.\n\n")

cat("2. P-VALUE DISTRIBUTION:\n")
cat("   Uniform distribution (flat histogram) = no biological signal or data issue.\n")
cat("   Enrichment near 0 = true biological differences.\n")
cat("   Peve's near-uniform distribution is concerning.\n\n")

cat("3. EXPRESSION VARIANCE:\n")
cat("   Lower CV = less variation = harder to detect differences.\n")
cat("   If Peve has systematically lower CV, this explains fewer findings.\n\n")

cat("4. PCA SEPARATION:\n")
cat("   Poor timepoint separation = weak temporal signal.\n")
cat("   This could be biological (stable expression) or technical.\n\n")

cat("5. RECOMMENDATIONS:\n")
cat("   - Check Peve alignment rates and read depth\n")
cat("   - Verify Peve genome/annotation quality\n")
cat("   - Consider whether Peve genuinely has stable expression\n")
cat("   - Look for batch effects in Peve samples\n")
```

## Visualization

```{r viz-as-counts, fig.width=10, fig.height=6}
# Reshape for plotting
as_summary_long <- as_summary %>%
  pivot_longer(
    cols = c(DTU_Candidates, Sig_Exons, Sig_Junctions),
    names_to = "Category",
    values_to = "Count"
  ) %>%
  mutate(Category = factor(Category, levels = c("DTU_Candidates", "Sig_Exons", "Sig_Junctions")))

# Bar plot
ggplot(as_summary_long, aes(x = Species, y = Count, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Alternative Splicing Events Across Species",
    subtitle = "Comparison of DTU, Exon, and Junction-level evidence",
    x = "Species",
    y = "Count",
    fill = "Evidence Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  ) +
  scale_fill_brewer(palette = "Set2")
```

# Save Results

```{r save-results}
# Save all results to output directory

# Apul
write_csv(apul_transcript_results, file.path(output_dir, "apul_transcript_DE_results.csv"))
write_csv(apul_dtu_candidates, file.path(output_dir, "apul_DTU_candidates.csv"))
write_csv(apul_exon_results, file.path(output_dir, "apul_exon_DE_results.csv"))
write_csv(apul_exon_sig, file.path(output_dir, "apul_significant_exons.csv"))
write_csv(apul_junction_results, file.path(output_dir, "apul_junction_DE_results.csv"))
write_csv(apul_junction_sig, file.path(output_dir, "apul_significant_junctions.csv"))
write_csv(apul_AS_summary, file.path(output_dir, "apul_AS_gene_summary.csv"))

# Peve
write_csv(peve_transcript_results, file.path(output_dir, "peve_transcript_DE_results.csv"))
write_csv(peve_dtu_candidates, file.path(output_dir, "peve_DTU_candidates.csv"))
write_csv(peve_exon_results, file.path(output_dir, "peve_exon_DE_results.csv"))
write_csv(peve_exon_sig, file.path(output_dir, "peve_significant_exons.csv"))
write_csv(peve_junction_results, file.path(output_dir, "peve_junction_DE_results.csv"))
write_csv(peve_junction_sig, file.path(output_dir, "peve_significant_junctions.csv"))
write_csv(peve_AS_summary, file.path(output_dir, "peve_AS_gene_summary.csv"))

# Ptua
write_csv(ptua_transcript_results, file.path(output_dir, "ptua_transcript_DE_results.csv"))
write_csv(ptua_dtu_candidates, file.path(output_dir, "ptua_DTU_candidates.csv"))
write_csv(ptua_exon_results, file.path(output_dir, "ptua_exon_DE_results.csv"))
write_csv(ptua_exon_sig, file.path(output_dir, "ptua_significant_exons.csv"))
write_csv(ptua_junction_results, file.path(output_dir, "ptua_junction_DE_results.csv"))
write_csv(ptua_junction_sig, file.path(output_dir, "ptua_significant_junctions.csv"))
write_csv(ptua_AS_summary, file.path(output_dir, "ptua_AS_gene_summary.csv"))

# Cross-species summary
write_csv(as_summary, file.path(output_dir, "cross_species_AS_summary.csv"))

cat("All results saved to:", output_dir, "\n")
```

# Session Info

```{r session-info}
sessionInfo()
```

# Notes

## Interpretation Guidelines

1. **DTU Candidates**: Genes where multiple transcripts are differentially expressed suggest condition-specific isoform switching.

2. **Significant Exons**: Exons significant independently of whole-gene expression indicate exon skipping or alternative exon events.

3. **Significant Junctions**: Differential intron usage suggests splice site switching events.

4. **Evidence Integration**: Genes with multiple types of evidence (DTU + exon + junction) are high-confidence AS candidates.

## Next Steps

Consider using **IsoformSwitchAnalyzeR** for:

- Visualization of isoform switches
- Prediction of functional consequences
- Open reading frame analysis
- Domain annotation changes

