---
title: "Hypothesis A: Epigenetic Lability of Ion-Transport Genes in Coral Calcification"
author: "Steven Roberts"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)
```

## Hypothesis Overview

**Main Hypothesis**: "Plastic vs buffered calcification is encoded by epigenetic lability of ion-transport genes"

Species differ in how epigenetically labile key calcification genes are:

- **Acropora pulchra**: High epigenetic responsiveness (dynamic DNA methylation) at calcification genes, enabling rapid changes but causing instability
- **Porites evermanni**: Epigenetic buffering (stable methylation, reduced regulatory turnover), enabling sustained calcification under stress

### "Frontloading" vs "Reacting" Hypothesis

- **Porites evermanni** exhibits constitutive high gene body methylation (GBM) in key calcification genes (Ca-ATPase, Carbonic Anhydrase), leading to stable, high-level transcription ("Frontloading")
- **Acropora pulchra** relies on promoter methylation changes to dynamically regulate these genes

### Expected Observations

1. Under stable conditions: Porites shows lower variance in methylation and expression compared to Acropora
2. Under heat stress (TP2): Acropora shows massive epigenetic remodeling (DMRs) associated with calcification collapse; Porites methylation profiles remain rigid

### Target Gene Systems

- Ca²⁺ transport: PMCAs, Ca²⁺ channels
- H⁺ removal / pH regulation: V-type ATPases, SLC transporters
- Carbon supply: carbonic anhydrases (CA)

---

## Setup and Libraries

```{r libraries}
library(tidyverse)
library(corrplot)
library(pheatmap)
library(RColorBrewer)
library(viridis)
library(ggpubr)
library(cowplot)
library(broom)

# Set theme
theme_set(theme_minimal(base_size = 12))

# Species lookup
species_lookup <- c(
  apul = "Acropora pulchra",
  peve = "Porites evermanni",
  ptua = "Pocillopora tuahiniensis"
)

# Color palette for species
species_colors <- c(
  "Acropora pulchra" = "#E64B35",
  "Porites evermanni" = "#4DBBD5",
  "Pocillopora tuahiniensis" = "#00A087"
)

# Timepoint colors
tp_colors <- c("TP1" = "#2166AC", "TP2" = "#D6604D", "TP3" = "#4393C3", "TP4" = "#92C5DE")
```

```{r output-dir}
output_dir <- "../output/51-HypA-epigenetic-lability"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

---

## Load Data

### Expression Matrices for Calcification Genes

```{r load-expression}
# Load biomineralization gene expression counts
apul_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/apul_biomin_counts.csv",
                        show_col_types = FALSE)

peve_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/peve_biomin_counts.csv",
                        show_col_types = FALSE)

ptua_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/ptua_biomin_counts.csv",
                        show_col_types = FALSE)
```

### Gene Body Methylation Data

```{r load-methylation}
# Acropora pulchra methylation
apul_meth <- read_tsv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/D-Apul/output/40-Apul-Gene-Methylation/Apul-gene-methylation_75pct.tsv",
                      show_col_types = FALSE)

# Porites evermanni methylation
peve_meth <- read_tsv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/E-Peve/output/15-Peve-Gene-Methylation/Peve-gene-methylation_75pct.tsv",
                      show_col_types = FALSE)

# Pocillopora tuahiniensis methylation
ptua_meth <- read_tsv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/F-Ptua/output/09-Ptua-Gene-Methylation/Ptua-gene-methylation_75pct.tsv",
                      show_col_types = FALSE)
```

### Annotation File

```{r load-annotation}
ortholog_annot <- read_csv("../output/12-ortho-annot/ortholog_groups_annotated.csv",
                           show_col_types = FALSE)
```

### Exon-Level Expression Data

```{r load-exon-data}
# Exon summary by ortholog for expression variability analysis
apul_exon_summary <- read_csv("../output/40-exon-count-matrix/apul-exon_summary_by_ortholog.csv",
                              show_col_types = FALSE)

peve_exon_summary <- read_csv("../output/40-exon-count-matrix/peve-exon_summary_by_ortholog.csv",
                              show_col_types = FALSE)

ptua_exon_summary <- read_csv("../output/40-exon-count-matrix/ptua-exon_summary_by_ortholog.csv",
                              show_col_types = FALSE)
```

---

## Data Wrangling

### Reshape Expression Data

```{r reshape-expression}
# Function to reshape expression data to long format
reshape_expression <- function(df, species_code) {
  df %>%
    pivot_longer(
      cols = -c(group_id, gene_id),
      names_to = "sample_id",
      values_to = "count"
    ) %>%
    mutate(
      species_code = species_code,
      colony_id = str_remove(sample_id, "-TP\\d+$"),
      timepoint = str_extract(sample_id, "TP\\d+"),
      timepoint_num = parse_number(timepoint),
      log_count = log2(count + 1)
    )
}

# Reshape all species
expr_long <- bind_rows(
  reshape_expression(apul_biomin, "apul"),
  reshape_expression(peve_biomin, "peve"),
  reshape_expression(ptua_biomin, "ptua")
) %>%
  mutate(species = species_lookup[species_code])

head(expr_long)
```

### Reshape Methylation Data

```{r reshape-methylation}
# Function to reshape methylation data
reshape_methylation <- function(df, species_code) {
  df %>%
    pivot_longer(
      cols = -gene_id,
      names_to = "sample_id",
      values_to = "methylation"
    ) %>%
    mutate(
      species_code = species_code,
      colony_id = str_remove(sample_id, "-TP\\d+$"),
      timepoint = str_extract(sample_id, "TP\\d+"),
      timepoint_num = parse_number(timepoint)
    )
}

# Reshape all species methylation
meth_long <- bind_rows(
  reshape_methylation(apul_meth, "apul") %>% mutate(gene_id = str_remove(gene_id, "^gene-")),
  reshape_methylation(peve_meth, "peve") %>% mutate(gene_id = str_remove(gene_id, "^gene-")),
  reshape_methylation(ptua_meth, "ptua") %>% mutate(gene_id = str_remove(gene_id, "^gene-"))
) %>%
  mutate(species = species_lookup[species_code])

head(meth_long)
```

### Identify Biomineralization Genes

```{r biomin-gene-ids}
# Get unique biomineralization gene IDs for each species
biomin_genes <- list(
  apul = unique(apul_biomin$gene_id),
  peve = unique(peve_biomin$gene_id),
  ptua = unique(ptua_biomin$gene_id)
)

biomin_orthogroups <- unique(c(apul_biomin$group_id, peve_biomin$group_id, ptua_biomin$group_id))

cat("Number of biomineralization genes per species:\n")
sapply(biomin_genes, length)
cat("\nTotal unique orthogroups:", length(biomin_orthogroups), "\n")
```

### Filter Methylation Data for Biomin Genes

```{r filter-biomin-methylation}
# Filter methylation data to biomineralization genes
biomin_meth <- meth_long %>%
  filter(
    (species_code == "apul" & gene_id %in% biomin_genes$apul) |
    (species_code == "peve" & gene_id %in% biomin_genes$peve) |
    (species_code == "ptua" & gene_id %in% biomin_genes$ptua)
  )

cat("Biomin genes with methylation data:\n")
biomin_meth %>%
  group_by(species_code) %>%
  summarise(n_genes = n_distinct(gene_id)) %>%
  print()
```

---
  
## Analysis 1: Methylation Variability Across Species

### Calculate Methylation Variance per Gene

```{r meth-variance}
# Calculate methylation variance for each gene within and across timepoints
meth_variance <- biomin_meth %>%
  filter(!is.na(methylation)) %>%
  group_by(species_code, species, gene_id) %>%
  summarise(
    mean_meth = mean(methylation, na.rm = TRUE),
    sd_meth = sd(methylation, na.rm = TRUE),
    cv_meth = sd_meth / mean_meth,  # Coefficient of variation
    n_samples = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(cv_meth) & is.finite(cv_meth))

# Summary by species
meth_variance_summary <- meth_variance %>%
  group_by(species_code, species) %>%
  summarise(
    mean_cv = mean(cv_meth, na.rm = TRUE),
    median_cv = median(cv_meth, na.rm = TRUE),
    sd_cv = sd(cv_meth, na.rm = TRUE),
    n_genes = n(),
    .groups = "drop"
  )

print(meth_variance_summary)
```

### Visualization: Methylation Variability Distribution

```{r meth-variance-plot, fig.height=6, fig.width=10}
# Boxplot of methylation CV by species
p1 <- ggplot(meth_variance, aes(x = species, y = cv_meth, fill = species)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.5) +
  scale_fill_manual(values = species_colors) +
  labs(
    title = "Methylation Variability (CV) Across Biomineralization Genes",
    subtitle = "Hypothesis: Acropora should show higher CV than Porites",
    x = "Species",
    y = "Coefficient of Variation (CV) in Methylation"
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(face = "italic"))

p1
ggsave(file.path(output_dir, "meth_cv_boxplot.png"), p1, width = 10, height = 6, dpi = 300)
```

### Statistical Test: Compare Methylation Variance

```{r meth-variance-test}
# Kruskal-Wallis test for methylation CV across species
kw_meth <- kruskal.test(cv_meth ~ species_code, data = meth_variance)
print(kw_meth)

# Pairwise Wilcoxon tests
pairwise_meth <- pairwise.wilcox.test(meth_variance$cv_meth, meth_variance$species_code,
                                       p.adjust.method = "bonferroni")
print(pairwise_meth)

# Save results
meth_test_results <- list(
  kruskal_wallis = kw_meth,
  pairwise_wilcox = pairwise_meth
)
```

---

## Analysis 2: Mean Gene Body Methylation Levels

### Compare Baseline GBM Levels

```{r baseline-gbm}
# Calculate mean methylation level per gene
mean_gbm <- biomin_meth %>%
  filter(!is.na(methylation)) %>%
  group_by(species_code, species, gene_id) %>%
  summarise(
    mean_meth = mean(methylation, na.rm = TRUE),
    .groups = "drop"
  )

# Summary by species
gbm_summary <- mean_gbm %>%
  group_by(species_code, species) %>%
  summarise(
    grand_mean = mean(mean_meth, na.rm = TRUE),
    median_meth = median(mean_meth, na.rm = TRUE),
    sd_meth = sd(mean_meth, na.rm = TRUE),
    n_genes = n(),
    .groups = "drop"
  )

print(gbm_summary)
```

### Visualization: Mean GBM Distribution

```{r gbm-distribution, fig.height=6, fig.width=10}
# Density plot of mean methylation by species
p2 <- ggplot(mean_gbm, aes(x = mean_meth, fill = species, color = species)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = species_colors) +
  scale_color_manual(values = species_colors) +
  coord_cartesian(ylim = c(0, .025)) +
  labs(
    title = "Distribution of Mean Gene Body Methylation in Biomineralization Genes",
    subtitle = "Hypothesis: Porites shows higher baseline GBM (Frontloading)",
    x = "Mean Methylation (%)",
    y = "Density"
  ) +
  theme(legend.position = "bottom",
        legend.text = element_text(face = "italic"))

p2
ggsave(file.path(output_dir, "gbm_density.png"), p2, width = 10, height = 6, dpi = 300)
```

### Statistical Test: Compare Mean GBM

```{r gbm-test}
# Kruskal-Wallis test for mean GBM across species
kw_gbm <- kruskal.test(mean_meth ~ species_code, data = mean_gbm)
print(kw_gbm)

# Pairwise Wilcoxon tests
pairwise_gbm <- pairwise.wilcox.test(mean_gbm$mean_meth, mean_gbm$species_code,
                                      p.adjust.method = "bonferroni")
print(pairwise_gbm)
```

---

## Analysis 3: Expression Variability Across Timepoints

### Calculate Expression Variance

```{r expr-variance}
# Calculate expression variance per gene
expr_variance <- expr_long %>%
  group_by(species_code, species, group_id, gene_id) %>%
  summarise(
    mean_count = mean(count, na.rm = TRUE),
    sd_count = sd(count, na.rm = TRUE),
    cv_expr = sd_count / mean_count,
    mean_log = mean(log_count, na.rm = TRUE),
    sd_log = sd(log_count, na.rm = TRUE),
    n_samples = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(cv_expr) & is.finite(cv_expr) & mean_count > 10)  # Filter low-expressed genes

# Summary by species
expr_var_summary <- expr_variance %>%
  group_by(species_code, species) %>%
  summarise(
    mean_cv = mean(cv_expr, na.rm = TRUE),
    median_cv = median(cv_expr, na.rm = TRUE),
    sd_cv = sd(cv_expr, na.rm = TRUE),
    n_genes = n(),
    .groups = "drop"
  )

print(expr_var_summary)
```

### Visualization: Expression Variability

```{r expr-variance-plot, fig.height=6, fig.width=10}
p3 <- ggplot(expr_variance, aes(x = species, y = cv_expr, fill = species)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.5) +
  scale_fill_manual(values = species_colors) +
  scale_y_continuous(limits = c(0, quantile(expr_variance$cv_expr, 0.95))) +
  labs(
    title = "Expression Variability (CV) in Biomineralization Genes",
    subtitle = "Hypothesis: Acropora shows higher transcriptional variance",
    x = "Species",
    y = "Coefficient of Variation (CV) in Expression"
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(face = "italic"))

p3
ggsave(file.path(output_dir, "expr_cv_boxplot.png"), p3, width = 10, height = 6, dpi = 300)
```

### Statistical Test: Compare Expression Variance

```{r expr-variance-test}
# Kruskal-Wallis test
kw_expr <- kruskal.test(cv_expr ~ species_code, data = expr_variance)
print(kw_expr)

# Pairwise tests
pairwise_expr <- pairwise.wilcox.test(expr_variance$cv_expr, expr_variance$species_code,
                                       p.adjust.method = "bonferroni")
print(pairwise_expr)
```

---

## Analysis 4: Temporal Dynamics - TP2 (Heat Stress) Response

### Calculate Timepoint-Specific Changes

```{r tp-dynamics}
# Calculate expression by timepoint
expr_by_tp <- expr_long %>%
  group_by(species_code, species, group_id, gene_id, timepoint, timepoint_num) %>%
  summarise(
    mean_count = mean(count, na.rm = TRUE),
    mean_log = mean(log_count, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate fold change relative to TP1
expr_fc <- expr_by_tp %>%
  group_by(species_code, species, group_id, gene_id) %>%
  mutate(
    tp1_mean = mean_log[timepoint == "TP1"],
    fc_from_tp1 = mean_log - tp1_mean
  ) %>%
  ungroup()
```

### Methylation by Timepoint

```{r meth-by-tp}
# Calculate methylation by timepoint
meth_by_tp <- biomin_meth %>%
  filter(!is.na(methylation)) %>%
  group_by(species_code, species, gene_id, timepoint, timepoint_num) %>%
  summarise(
    mean_meth = mean(methylation, na.rm = TRUE),
    sd_meth = sd(methylation, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate methylation change from TP1
meth_delta <- meth_by_tp %>%
  group_by(species_code, species, gene_id) %>%
  mutate(
    tp1_meth = mean_meth[timepoint == "TP1"],
    delta_meth = mean_meth - tp1_meth
  ) %>%
  ungroup() %>%
  filter(!is.na(delta_meth))
```

### Visualization: Methylation Changes Across Timepoints

```{r meth-dynamics-plot, fig.height=8, fig.width=12}
# Boxplot of methylation change from TP1
p4 <- meth_delta %>%
  filter(timepoint != "TP1") %>%
  ggplot(aes(x = timepoint, y = delta_meth, fill = species)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = species_colors) +
  facet_wrap(~species, ncol = 3) +
  labs(
    title = "Methylation Changes Relative to TP1 (Baseline)",
    subtitle = "TP2 = Heat Stress | Hypothesis: Acropora shows larger remodeling at TP2",
    x = "Timepoint",
    y = "Δ Methylation (from TP1)"
  ) +
  theme(legend.position = "none",
        strip.text = element_text(face = "italic"))

p4
ggsave(file.path(output_dir, "meth_delta_tp.png"), p4, width = 12, height = 8, dpi = 300)
```

### Quantify Remodeling Magnitude

```{r remodeling-magnitude}
# Calculate absolute methylation change magnitude at each timepoint
remodeling_magnitude <- meth_delta %>%
  filter(timepoint != "TP1") %>%
  group_by(species_code, species, timepoint) %>%
  summarise(
    mean_abs_delta = mean(abs(delta_meth), na.rm = TRUE),
    median_abs_delta = median(abs(delta_meth), na.rm = TRUE),
    n_genes = n(),
    .groups = "drop"
  )

print(remodeling_magnitude)

# Visualization
p5 <- ggplot(remodeling_magnitude, aes(x = timepoint, y = mean_abs_delta, 
                                       fill = species, group = species)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), alpha = 0.8) +
  scale_fill_manual(values = species_colors) +
  labs(
    title = "Magnitude of Methylation Remodeling by Timepoint",
    subtitle = "Higher values = more epigenetic remodeling",
    x = "Timepoint",
    y = "Mean |Δ Methylation|"
  ) +
  theme(legend.text = element_text(face = "italic"))

p5
ggsave(file.path(output_dir, "remodeling_magnitude.png"), p5, width = 10, height = 6, dpi = 300)
```

---

## Analysis 5: Expression-Methylation Correlation

### Merge Expression and Methylation Data

```{r merge-expr-meth}
# Create gene mapping between expression and methylation
# Expression uses gene_id from biomin counts
# Methylation uses gene_id from methylation files

# Prepare expression summary by gene
expr_summary <- expr_long %>%
  group_by(species_code, gene_id) %>%
  summarise(
    mean_expr = mean(log_count, na.rm = TRUE),
    cv_expr = sd(log_count, na.rm = TRUE) / mean(log_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(cv_expr) & is.finite(cv_expr))

# Merge with methylation variance
expr_meth_combined <- meth_variance %>%
  inner_join(expr_summary, by = c("species_code", "gene_id"))

cat("Genes with both expression and methylation data:\n")
expr_meth_combined %>%
  group_by(species_code) %>%
  summarise(n_genes = n()) %>%
  print()
```

### Correlation Analysis

```{r expr-meth-correlation}
# Calculate correlation between methylation and expression CV
correlations <- expr_meth_combined %>%
  group_by(species_code, species) %>%
  summarise(
    cor_mean_meth_expr = cor(mean_meth, mean_expr, use = "complete.obs"),
    cor_cv_meth_cv_expr = cor(cv_meth, cv_expr, use = "complete.obs", method = "spearman"),
    n = n(),
    .groups = "drop"
  )

print(correlations)
```

### Visualization: Expression vs Methylation

```{r expr-meth-scatter, fig.height=8, fig.width=12}
# Scatter plot: Mean methylation vs Mean expression
p6 <- ggplot(expr_meth_combined, aes(x = mean_meth, y = mean_expr, color = species)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  scale_color_manual(values = species_colors) +
  facet_wrap(~species, scales = "free") +
  labs(
    title = "Gene Body Methylation vs Expression Level",
    subtitle = "Biomineralization genes",
    x = "Mean Methylation (%)",
    y = "Mean Expression (log2)"
  ) +
  theme(legend.position = "none",
        strip.text = element_text(face = "italic"))

p6
ggsave(file.path(output_dir, "expr_meth_scatter.png"), p6, width = 12, height = 8, dpi = 300)
```

### Variability Correlation

```{r cv-correlation, fig.height=8, fig.width=12}
# Scatter plot: CV methylation vs CV expression
p7 <- ggplot(expr_meth_combined, aes(x = cv_meth, y = cv_expr, color = species)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  scale_color_manual(values = species_colors) +
  scale_x_continuous(limits = c(0, quantile(expr_meth_combined$cv_meth, 0.95, na.rm = TRUE))) +
  scale_y_continuous(limits = c(0, quantile(expr_meth_combined$cv_expr, 0.95, na.rm = TRUE))) +
  facet_wrap(~species, scales = "free") +
  labs(
    title = "Methylation Variability vs Expression Variability",
    subtitle = "Higher methylation CV may indicate epigenetic lability",
    x = "CV Methylation",
    y = "CV Expression"
  ) +
  theme(legend.position = "none",
        strip.text = element_text(face = "italic"))

p7
ggsave(file.path(output_dir, "cv_meth_expr_scatter.png"), p7, width = 12, height = 8, dpi = 300)
```

---

## Analysis 6: Target Gene System Analysis

### Annotate Target Gene Systems

```{r target-genes}
# Define target gene keywords
target_keywords <- list(
  calcium_transport = c("PMCA", "plasma membrane calcium", "calcium channel", 
                        "calcium-transporting", "Ca2+-ATPase", "SERCA", "calcium pump"),
  ph_regulation = c("V-type ATPase", "vacuolar ATPase", "SLC", "solute carrier", 
                    "sodium/hydrogen", "Na+/H+", "NHE", "proton pump"),
  carbon_supply = c("carbonic anhydrase", "CA", "bicarbonate")
)

# Search for target genes in annotation
find_target_genes <- function(annotations, keywords) {
  pattern <- paste(keywords, collapse = "|")
  annotations %>%
    filter(str_detect(protein_name, regex(pattern, ignore_case = TRUE)) |
           str_detect(go_mf, regex(pattern, ignore_case = TRUE)) |
           str_detect(go_bp, regex(pattern, ignore_case = TRUE)))
}

# Filter annotation to biomineralization orthogroups
biomin_annot <- ortholog_annot %>%
  filter(group_id %in% biomin_orthogroups)

# Find target genes
target_genes <- list(
  calcium = find_target_genes(biomin_annot, target_keywords$calcium_transport),
  ph = find_target_genes(biomin_annot, target_keywords$ph_regulation),
  carbon = find_target_genes(biomin_annot, target_keywords$carbon_supply)
)

cat("Target gene counts:\n")
cat("  Calcium transport:", nrow(target_genes$calcium), "\n")
cat("  pH regulation:", nrow(target_genes$ph), "\n")
cat("  Carbon supply:", nrow(target_genes$carbon), "\n")
```

### Focused Analysis on Target Genes

```{r target-gene-analysis}
# Combine target gene orthogroups
target_ogs <- unique(c(
  target_genes$calcium$group_id,
  target_genes$ph$group_id,
  target_genes$carbon$group_id
))

cat("Total target orthogroups:", length(target_ogs), "\n")

# Filter variance data to target genes
target_meth_var <- meth_variance %>%
  inner_join(
    bind_rows(
      apul_biomin %>% select(group_id, gene_id),
      peve_biomin %>% select(group_id, gene_id),
      ptua_biomin %>% select(group_id, gene_id)
    ) %>% filter(group_id %in% target_ogs),
    by = "gene_id"
  )

target_expr_var <- expr_variance %>%
  filter(group_id %in% target_ogs)
```

### Visualization: Target Gene Variability

```{r target-gene-plots, fig.height=6, fig.width=12}
if(nrow(target_meth_var) > 0) {
  # Methylation CV for target genes
  p8a <- ggplot(target_meth_var, aes(x = species, y = cv_meth, fill = species)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5) +
    scale_fill_manual(values = species_colors) +
    labs(
      title = "Methylation Variability in Target Calcification Genes",
      subtitle = "Ca-ATPases, V-ATPases, Carbonic Anhydrases",
      x = "Species",
      y = "CV Methylation"
    ) +
    theme(legend.position = "none",
          axis.text.x = element_text(face = "italic"))
  
  print(p8a)
  ggsave(file.path(output_dir, "target_genes_meth_cv.png"), p8a, width = 10, height = 6, dpi = 300)
}

if(nrow(target_expr_var) > 0) {
  # Expression CV for target genes
  p8b <- ggplot(target_expr_var, aes(x = species, y = cv_expr, fill = species)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5) +
    scale_fill_manual(values = species_colors) +
    labs(
      title = "Expression Variability in Target Calcification Genes",
      x = "Species",
      y = "CV Expression"
    ) +
    theme(legend.position = "none",
          axis.text.x = element_text(face = "italic"))
  
  print(p8b)
  ggsave(file.path(output_dir, "target_genes_expr_cv.png"), p8b, width = 10, height = 6, dpi = 300)
}
```

---

## Analysis 7: Exon-Level Expression Variability (Alternative Splicing Proxy)

### Load and Analyze Exon Data

```{r exon-analysis}
# Combine exon summaries
exon_combined <- bind_rows(
  apul_exon_summary %>% mutate(species_code = "apul"),
  peve_exon_summary %>% mutate(species_code = "peve"),
  ptua_exon_summary %>% mutate(species_code = "ptua")
) %>%
  mutate(species = species_lookup[species_code])

# Filter to biomineralization genes
biomin_exon <- exon_combined %>%
  filter(group_id %in% biomin_orthogroups | gene_id %in% unlist(biomin_genes))

# Summary statistics
exon_var_summary <- biomin_exon %>%
  group_by(species_code, species) %>%
  summarise(
    mean_cv = mean(cv_expr, na.rm = TRUE),
    median_cv = median(cv_expr, na.rm = TRUE),
    n_exons = n(),
    .groups = "drop"
  )

print(exon_var_summary)
```

### Visualization: Exon-Level Variability

```{r exon-variance-plot, fig.height=6, fig.width=10}
if(nrow(biomin_exon) > 0) {
  # Filter out extreme outliers for visualization
  exon_plot_data <- biomin_exon %>%
    filter(cv_expr < quantile(cv_expr, 0.95, na.rm = TRUE))
  
  p9 <- ggplot(exon_plot_data, aes(x = species, y = cv_expr, fill = species)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.2, alpha = 0.8) +
    scale_fill_manual(values = species_colors) +
    labs(
      title = "Exon-Level Expression Variability in Biomineralization Genes",
      subtitle = "Proxy for alternative splicing / regulatory complexity",
      x = "Species",
      y = "CV Expression (Exon-Level)"
    ) +
    theme(legend.position = "none",
          axis.text.x = element_text(face = "italic"))
  
  print(p9)
  ggsave(file.path(output_dir, "exon_cv_violin.png"), p9, width = 10, height = 6, dpi = 300)
}
```

---

## Summary and Hypothesis Evaluation

### Compile Results Summary

```{r results-summary}
# Create summary table
summary_table <- tibble(
  Metric = c(
    "Mean Methylation CV",
    "Mean GBM Level",
    "Mean Expression CV",
    "Remodeling Magnitude (TP2)"
  )
)

# Add species values
for(sp in c("apul", "peve", "ptua")) {
  summary_table[[species_lookup[sp]]] <- c(
    meth_variance_summary %>% filter(species_code == sp) %>% pull(mean_cv) %>% round(3),
    gbm_summary %>% filter(species_code == sp) %>% pull(grand_mean) %>% round(2),
    expr_var_summary %>% filter(species_code == sp) %>% pull(mean_cv) %>% round(3),
    remodeling_magnitude %>% filter(species_code == sp, timepoint == "TP2") %>% 
      pull(mean_abs_delta) %>% round(3)
  )
}

print(summary_table)

# Save summary
write_csv(summary_table, file.path(output_dir, "hypothesis_summary.csv"))
```

### Hypothesis Evaluation

```{r hypothesis-evaluation}
cat("\n=== HYPOTHESIS EVALUATION ===\n\n")

cat("Hypothesis 1: Acropora shows higher methylation variability (CV) than Porites\n")
apul_cv <- meth_variance_summary %>% filter(species_code == "apul") %>% pull(mean_cv)
peve_cv <- meth_variance_summary %>% filter(species_code == "peve") %>% pull(mean_cv)
cat(sprintf("  Acropora CV: %.3f | Porites CV: %.3f\n", apul_cv, peve_cv))
cat(sprintf("  Result: %s\n\n", ifelse(apul_cv > peve_cv, "SUPPORTED", "NOT SUPPORTED")))

cat("Hypothesis 2: Porites has higher baseline GBM (Frontloading)\n")
apul_gbm <- gbm_summary %>% filter(species_code == "apul") %>% pull(grand_mean)
peve_gbm <- gbm_summary %>% filter(species_code == "peve") %>% pull(grand_mean)
cat(sprintf("  Acropora GBM: %.2f | Porites GBM: %.2f\n", apul_gbm, peve_gbm))
cat(sprintf("  Result: %s\n\n", ifelse(peve_gbm > apul_gbm, "SUPPORTED", "NOT SUPPORTED")))

cat("Hypothesis 3: Acropora shows higher expression variability\n")
apul_expr_cv <- expr_var_summary %>% filter(species_code == "apul") %>% pull(mean_cv)
peve_expr_cv <- expr_var_summary %>% filter(species_code == "peve") %>% pull(mean_cv)
cat(sprintf("  Acropora Expr CV: %.3f | Porites Expr CV: %.3f\n", apul_expr_cv, peve_expr_cv))
cat(sprintf("  Result: %s\n\n", ifelse(apul_expr_cv > peve_expr_cv, "SUPPORTED", "NOT SUPPORTED")))

cat("Hypothesis 4: Acropora shows greater methylation remodeling at TP2 (heat stress)\n")
apul_tp2 <- remodeling_magnitude %>% filter(species_code == "apul", timepoint == "TP2") %>% pull(mean_abs_delta)
peve_tp2 <- remodeling_magnitude %>% filter(species_code == "peve", timepoint == "TP2") %>% pull(mean_abs_delta)
cat(sprintf("  Acropora TP2 |Δmeth|: %.3f | Porites TP2 |Δmeth|: %.3f\n", apul_tp2, peve_tp2))
cat(sprintf("  Result: %s\n", ifelse(apul_tp2 > peve_tp2, "SUPPORTED", "NOT SUPPORTED")))
```

---

## Combined Visualization

```{r combined-plot, fig.height=12, fig.width=14}
# Create combined summary figure
combined_plot <- plot_grid(
  p1 + theme(plot.title = element_text(size = 10)),
  p2 + theme(plot.title = element_text(size = 10)),
  p3 + theme(plot.title = element_text(size = 10)),
  p5 + theme(plot.title = element_text(size = 10)),
  ncol = 2,
  labels = c("A", "B", "C", "D")
)

combined_plot
ggsave(file.path(output_dir, "combined_hypothesis_figure.png"), 
       combined_plot, width = 14, height = 12, dpi = 300)
```

---

## Session Info

```{r session-info}
sessionInfo()
```

