---
title: "Calcification Gene Expression Variability Across Coral Species"
author: "Timeseries Molecular Calcification"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = TRUE
)
```

# Introduction

This analysis tests the hypothesis that **gene expression of calcification genes is more variable in *Acropora* and *Pocillopora* than *Porites***.

To assess variability, we examine expression across samples and time, considering both:
1. Gene-level expression variability
2. Exon-level expression variability (alternative splicing patterns)

# Load Libraries

```{r libraries}
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(ggpubr)
library(pheatmap)
library(viridis)
library(reshape2)
library(scales)
library(knitr)
library(kableExtra)
```

# Data Loading

## Biomin Expression Counts

Expression matrices of calcification genes across timepoints and treatments for each species.

```{r load-biomin-counts}
# Load biomin count data from GitHub
apul_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/apul_biomin_counts.csv")
peve_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/peve_biomin_counts.csv")
ptua_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/ptua_biomin_counts.csv")

# Preview the data
cat("Apul biomin genes:", nrow(apul_biomin), "\n")
cat("Peve biomin genes:", nrow(peve_biomin), "\n")
cat("Ptua biomin genes:", nrow(ptua_biomin), "\n")
```

## Annotation File

```{r load-annotations}
# Load ortholog annotations
ortho_annot <- read_csv("../output/12-ortho-annot/ortholog_groups_annotated.csv")

# Preview structure
glimpse(ortho_annot)
```

## Exon-Level Expression Data

```{r load-exon-data}
# Load exon summary files with ortholog group mappings
# These files contain pre-computed variability metrics per exon

apul_exon_summary <- read_csv("../output/40-exon-count-matrix/apul-exon_summary_by_ortholog.csv")
peve_exon_summary <- read_csv("../output/40-exon-count-matrix/peve-exon_summary_by_ortholog.csv")
ptua_exon_summary <- read_csv("../output/40-exon-count-matrix/ptua-exon_summary_by_ortholog.csv")

cat("Apul exons:", nrow(apul_exon_summary), "\n")
cat("Peve exons:", nrow(peve_exon_summary), "\n")
cat("Ptua exons:", nrow(ptua_exon_summary), "\n")
```

# Data Preprocessing

## Extract Sample Metadata from Column Names

```{r extract-metadata}
# Function to extract sample info from column names
extract_sample_info <- function(counts_df, species_prefix) {
  sample_cols <- colnames(counts_df)[!colnames(counts_df) %in% c("group_id", "gene_id")]
  
  sample_info <- data.frame(
    sample_id = sample_cols,
    stringsAsFactors = FALSE
  ) %>%
    mutate(
      # Extract colony ID and timepoint
      parts = str_match(sample_id, paste0(species_prefix, "-(\\d+)-TP(\\d+)")),
      colony = parts[,2],
      timepoint = paste0("TP", parts[,3]),
      species = species_prefix
    ) %>%
    select(sample_id, species, colony, timepoint)
  
  return(sample_info)
}

# Extract metadata for each species
apul_meta <- extract_sample_info(apul_biomin, "ACR")
peve_meta <- extract_sample_info(peve_biomin, "POR")
ptua_meta <- extract_sample_info(ptua_biomin, "POC")

# Combine metadata
all_meta <- bind_rows(apul_meta, peve_meta, ptua_meta) %>%
  mutate(species_full = case_when(
    species == "ACR" ~ "Acropora pulchra",
    species == "POR" ~ "Porites evermanni",
    species == "POC" ~ "Pocillopora tuahiniensis"
  ))

kable(head(all_meta, 10)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Get Common Ortholog Groups

```{r common-orthologs}
# Find ortholog groups present in all three species' biomin datasets
apul_groups <- unique(apul_biomin$group_id)
peve_groups <- unique(peve_biomin$group_id)
ptua_groups <- unique(ptua_biomin$group_id)

# Common calcification-related ortholog groups
common_groups <- Reduce(intersect, list(apul_groups, peve_groups, ptua_groups))

cat("Apul biomin ortholog groups:", length(apul_groups), "\n")
cat("Peve biomin ortholog groups:", length(peve_groups), "\n")
cat("Ptua biomin ortholog groups:", length(ptua_groups), "\n")
cat("Common ortholog groups:", length(common_groups), "\n")
```

# Gene-Level Expression Variability Analysis

## Calculate Coefficient of Variation (CV) for Each Gene

```{r calculate-gene-cv}
# Function to calculate CV for each gene across all samples
calc_gene_cv <- function(counts_df, species_name) {
  # Get sample columns (exclude group_id and gene_id)
  sample_cols <- colnames(counts_df)[!colnames(counts_df) %in% c("group_id", "gene_id")]
  
  counts_matrix <- counts_df %>%
    select(all_of(sample_cols)) %>%
    as.matrix()
  
  # Calculate mean, SD, and CV for each gene
  gene_stats <- data.frame(
    group_id = counts_df$group_id,
    gene_id = counts_df$gene_id,
    mean_expr = rowMeans(counts_matrix, na.rm = TRUE),
    sd_expr = apply(counts_matrix, 1, sd, na.rm = TRUE),
    n_samples = ncol(counts_matrix),
    species = species_name
  ) %>%
    mutate(
      cv_expr = sd_expr / mean_expr,
      # Handle cases where mean is 0
      cv_expr = ifelse(is.infinite(cv_expr) | is.nan(cv_expr), NA, cv_expr)
    )
  
  return(gene_stats)
}

# Calculate CV for each species
apul_cv <- calc_gene_cv(apul_biomin, "Acropora")
peve_cv <- calc_gene_cv(peve_biomin, "Porites")
ptua_cv <- calc_gene_cv(ptua_biomin, "Pocillopora")

# Combine all species
all_gene_cv <- bind_rows(apul_cv, peve_cv, ptua_cv) %>%
  # Filter for expressed genes (mean > 1)
  filter(mean_expr > 1)

cat("Genes with expression > 1:\n")
cat("  Acropora:", sum(all_gene_cv$species == "Acropora"), "\n")
cat("  Porites:", sum(all_gene_cv$species == "Porites"), "\n")
cat("  Pocillopora:", sum(all_gene_cv$species == "Pocillopora"), "\n")
```

## Compare CV Distributions Across Species

```{r cv-distribution-comparison, fig.height=8, fig.width=10}
# Summary statistics by species
cv_summary <- all_gene_cv %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    mean_cv = mean(cv_expr, na.rm = TRUE),
    median_cv = median(cv_expr, na.rm = TRUE),
    sd_cv = sd(cv_expr, na.rm = TRUE),
    iqr_cv = IQR(cv_expr, na.rm = TRUE),
    q25 = quantile(cv_expr, 0.25, na.rm = TRUE),
    q75 = quantile(cv_expr, 0.75, na.rm = TRUE)
  )

kable(cv_summary, digits = 3, caption = "Gene Expression CV Summary by Species") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Violin plot of CV distribution
p1 <- ggplot(all_gene_cv, aes(x = species, y = cv_expr, fill = species)) +
  geom_violin(alpha = 0.7, trim = TRUE) +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
  scale_fill_manual(values = c("Acropora" = "#E64B35", 
                                "Porites" = "#4DBBD5", 
                                "Pocillopora" = "#00A087")) +
  labs(
    title = "Gene Expression Variability (CV) in Calcification Genes",
    subtitle = "Higher CV indicates more variable expression across samples",
    x = "Species",
    y = "Coefficient of Variation (CV)"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

print(p1)
```

## Statistical Tests for CV Differences

```{r cv-statistical-tests}
# Kruskal-Wallis test for overall differences
kw_test <- kruskal.test(cv_expr ~ species, data = all_gene_cv)
cat("Kruskal-Wallis test for CV differences:\n")
print(kw_test)

# Pairwise Wilcoxon tests
pairwise_tests <- pairwise.wilcox.test(
  all_gene_cv$cv_expr, 
  all_gene_cv$species, 
  p.adjust.method = "bonferroni"
)
cat("\nPairwise Wilcoxon tests (Bonferroni adjusted):\n")
print(pairwise_tests)

# Compare specific pairs
acro_vs_porites <- wilcox.test(
  all_gene_cv$cv_expr[all_gene_cv$species == "Acropora"],
  all_gene_cv$cv_expr[all_gene_cv$species == "Porites"]
)

pocillo_vs_porites <- wilcox.test(
  all_gene_cv$cv_expr[all_gene_cv$species == "Pocillopora"],
  all_gene_cv$cv_expr[all_gene_cv$species == "Porites"]
)

cat("\nAcropora vs Porites p-value:", acro_vs_porites$p.value, "\n")
cat("Pocillopora vs Porites p-value:", pocillo_vs_porites$p.value, "\n")
```

## CV Comparison for Common Ortholog Groups

```{r cv-common-orthologs, fig.height=8, fig.width=12}
# Filter for common ortholog groups only
common_cv <- all_gene_cv %>%
  filter(group_id %in% common_groups)

cat("Common ortholog genes with expression > 1:\n")
cat("  Acropora:", sum(common_cv$species == "Acropora"), "\n")
cat("  Porites:", sum(common_cv$species == "Porites"), "\n")
cat("  Pocillopora:", sum(common_cv$species == "Pocillopora"), "\n")

# Summary for common orthologs
common_cv_summary <- common_cv %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    mean_cv = mean(cv_expr, na.rm = TRUE),
    median_cv = median(cv_expr, na.rm = TRUE),
    sd_cv = sd(cv_expr, na.rm = TRUE)
  )

kable(common_cv_summary, digits = 3, 
      caption = "CV Summary for Common Calcification Orthologs") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Paired comparison plot for common orthologs
common_wide <- common_cv %>%
  select(group_id, species, cv_expr) %>%
  pivot_wider(names_from = species, values_from = cv_expr) %>%
  drop_na()

p2 <- ggplot(common_wide) +
  geom_point(aes(x = Porites, y = Acropora), alpha = 0.6, color = "#E64B35") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  labs(
    title = "Acropora vs Porites CV (Common Orthologs)",
    subtitle = paste("Points above line: higher variability in Acropora\n",
                     "n =", nrow(common_wide), "genes"),
    x = "Porites CV",
    y = "Acropora CV"
  ) +
  theme_minimal() +
  coord_equal()

p3 <- ggplot(common_wide) +
  geom_point(aes(x = Porites, y = Pocillopora), alpha = 0.6, color = "#00A087") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  labs(
    title = "Pocillopora vs Porites CV (Common Orthologs)",
    subtitle = paste("Points above line: higher variability in Pocillopora\n",
                     "n =", nrow(common_wide), "genes"),
    x = "Porites CV",
    y = "Pocillopora CV"
  ) +
  theme_minimal() +
  coord_equal()

ggarrange(p2, p3, ncol = 2)
```

# Temporal Variability Analysis

## Calculate Within-Colony and Between-Colony Variability

```{r temporal-variability}
# Function to calculate temporal variability components
calc_temporal_variability <- function(counts_df, meta_df, species_name) {
  sample_cols <- colnames(counts_df)[!colnames(counts_df) %in% c("group_id", "gene_id")]
  
  results <- counts_df %>%
    pivot_longer(cols = all_of(sample_cols), 
                 names_to = "sample_id", 
                 values_to = "count") %>%
    left_join(meta_df, by = "sample_id") %>%
    group_by(group_id, gene_id, colony) %>%
    summarise(
      within_colony_cv = sd(count, na.rm = TRUE) / mean(count, na.rm = TRUE),
      mean_count = mean(count, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    group_by(group_id, gene_id) %>%
    summarise(
      mean_within_colony_cv = mean(within_colony_cv, na.rm = TRUE),
      between_colony_cv = sd(mean_count, na.rm = TRUE) / mean(mean_count, na.rm = TRUE),
      overall_mean = mean(mean_count, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      species = species_name,
      mean_within_colony_cv = ifelse(is.infinite(mean_within_colony_cv) | is.nan(mean_within_colony_cv), 
                                     NA, mean_within_colony_cv),
      between_colony_cv = ifelse(is.infinite(between_colony_cv) | is.nan(between_colony_cv), 
                                 NA, between_colony_cv)
    )
  
  return(results)
}

# Calculate temporal variability
apul_temporal <- calc_temporal_variability(apul_biomin, apul_meta, "Acropora")
peve_temporal <- calc_temporal_variability(peve_biomin, peve_meta, "Porites")
ptua_temporal <- calc_temporal_variability(ptua_biomin, ptua_meta, "Pocillopora")

# Combine all
all_temporal <- bind_rows(apul_temporal, peve_temporal, ptua_temporal) %>%
  filter(overall_mean > 1)  # Filter for expressed genes

# Summary
temporal_summary <- all_temporal %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    mean_within_cv = mean(mean_within_colony_cv, na.rm = TRUE),
    mean_between_cv = mean(between_colony_cv, na.rm = TRUE),
    median_within_cv = median(mean_within_colony_cv, na.rm = TRUE),
    median_between_cv = median(between_colony_cv, na.rm = TRUE)
  )

kable(temporal_summary, digits = 3, 
      caption = "Temporal Variability Summary (Within vs Between Colony)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Temporal Variability

```{r temporal-plots, fig.height=8, fig.width=12}
# Reshape for plotting
temporal_long <- all_temporal %>%
  pivot_longer(
    cols = c(mean_within_colony_cv, between_colony_cv),
    names_to = "variability_type",
    values_to = "cv"
  ) %>%
  mutate(
    variability_type = ifelse(variability_type == "mean_within_colony_cv",
                              "Within-Colony (Temporal)", "Between-Colony")
  )

p4 <- ggplot(temporal_long, aes(x = species, y = cv, fill = variability_type)) +
  geom_violin(alpha = 0.7, position = position_dodge(0.8), trim = TRUE) +
  geom_boxplot(width = 0.15, position = position_dodge(0.8), 
               fill = "white", outlier.size = 0.5) +
  scale_fill_manual(values = c("Within-Colony (Temporal)" = "#7570B3", 
                                "Between-Colony" = "#E7298A")) +
  labs(
    title = "Within-Colony (Temporal) vs Between-Colony Variability",
    subtitle = "Calcification gene expression in coral species",
    x = "Species",
    y = "Coefficient of Variation (CV)",
    fill = "Variability Type"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

print(p4)
```

# Exon-Level Expression Variability

## Filter Exon Data for Calcification Genes

```{r filter-exon-biomin}
# Get the group_ids from biomin datasets
biomin_groups_apul <- unique(apul_biomin$group_id)
biomin_groups_peve <- unique(peve_biomin$group_id)
biomin_groups_ptua <- unique(ptua_biomin$group_id)

# Filter exon summary data for calcification-related genes using group_id
apul_exon_biomin <- apul_exon_summary %>%
  filter(group_id %in% biomin_groups_apul) %>%
  mutate(species = "Acropora")

peve_exon_biomin <- peve_exon_summary %>%
  filter(group_id %in% biomin_groups_peve) %>%
  mutate(species = "Porites")

ptua_exon_biomin <- ptua_exon_summary %>%
  filter(group_id %in% biomin_groups_ptua) %>%
  mutate(species = "Pocillopora")

cat("Exons in calcification genes:\n")
cat("  Acropora:", nrow(apul_exon_biomin), "\n")
cat("  Porites:", nrow(peve_exon_biomin), "\n")
cat("  Pocillopora:", nrow(ptua_exon_biomin), "\n")
```

## Compare Exon-Level CV Across Species

```{r exon-cv-comparison, fig.height=8, fig.width=10}
# Combine exon data
all_exon_cv <- bind_rows(apul_exon_biomin, peve_exon_biomin, ptua_exon_biomin) %>%
  filter(mean_expr > 0.5)  # Filter for expressed exons

# Summary statistics by species
exon_cv_summary <- all_exon_cv %>%
  group_by(species) %>%
  summarise(
    n_exons = n(),
    n_genes = n_distinct(group_id, na.rm = TRUE),
    mean_cv = mean(cv_expr, na.rm = TRUE),
    median_cv = median(cv_expr, na.rm = TRUE),
    sd_cv = sd(cv_expr, na.rm = TRUE)
  )

kable(exon_cv_summary, digits = 3, 
      caption = "Exon-Level Expression CV Summary by Species") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Violin plot of exon CV distribution
p5 <- ggplot(all_exon_cv, aes(x = species, y = cv_expr, fill = species)) +
  geom_violin(alpha = 0.7, trim = TRUE) +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
  scale_fill_manual(values = c("Acropora" = "#E64B35", 
                                "Porites" = "#4DBBD5", 
                                "Pocillopora" = "#00A087")) +
  labs(
    title = "Exon-Level Expression Variability (CV) in Calcification Genes",
    subtitle = "Higher CV indicates more variable exon expression across samples",
    x = "Species",
    y = "Coefficient of Variation (CV)"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

print(p5)
```

## Statistical Tests for Exon CV

```{r exon-cv-stats}
# Kruskal-Wallis test
kw_exon <- kruskal.test(cv_expr ~ species, data = all_exon_cv)
cat("Kruskal-Wallis test for Exon CV differences:\n")
print(kw_exon)

# Pairwise tests
pairwise_exon <- pairwise.wilcox.test(
  all_exon_cv$cv_expr, 
  all_exon_cv$species, 
  p.adjust.method = "bonferroni"
)
cat("\nPairwise Wilcoxon tests for Exon CV (Bonferroni adjusted):\n")
print(pairwise_exon)
```

## Exon Variability Within Genes (Alternative Splicing Proxy)

```{r exon-within-gene-variability}
# Calculate within-gene exon variability (proxy for alternative splicing)
exon_within_gene <- all_exon_cv %>%
  group_by(species, group_id) %>%
  summarise(
    n_exons = n(),
    exon_cv_mean = mean(cv_expr, na.rm = TRUE),
    exon_cv_sd = sd(cv_expr, na.rm = TRUE),
    exon_cv_range = max(cv_expr, na.rm = TRUE) - min(cv_expr, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_exons >= 2)  # Need at least 2 exons

# Summary by species
within_gene_summary <- exon_within_gene %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    mean_exon_cv_range = mean(exon_cv_range, na.rm = TRUE),
    median_exon_cv_range = median(exon_cv_range, na.rm = TRUE),
    mean_exon_cv_sd = mean(exon_cv_sd, na.rm = TRUE)
  )

kable(within_gene_summary, digits = 3, 
      caption = "Within-Gene Exon CV Variability (Alternative Splicing Proxy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Plot within-gene variability
p6 <- ggplot(exon_within_gene, aes(x = species, y = exon_cv_range, fill = species)) +
  geom_violin(alpha = 0.7, trim = TRUE) +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
  scale_fill_manual(values = c("Acropora" = "#E64B35", 
                                "Porites" = "#4DBBD5", 
                                "Pocillopora" = "#00A087")) +
  labs(
    title = "Within-Gene Exon CV Range (Alternative Splicing Indicator)",
    subtitle = "Higher range indicates more variable exon usage patterns",
    x = "Species",
    y = "CV Range (Max - Min) Across Exons"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

print(p6)

# Statistical test
kw_within <- kruskal.test(exon_cv_range ~ species, data = exon_within_gene)
cat("\nKruskal-Wallis test for within-gene exon variability:\n")
print(kw_within)
```

# Highly Variable Genes Analysis

## Identify Top Variable Genes Per Species

```{r top-variable-genes}
# Get top 20 most variable genes per species
top_variable <- all_gene_cv %>%
  group_by(species) %>%
  slice_max(order_by = cv_expr, n = 20) %>%
  left_join(ortho_annot %>% select(group_id, protein_name), by = "group_id") %>%
  select(species, group_id, gene_id, mean_expr, cv_expr, protein_name)

# Display top variable genes
kable(top_variable %>% 
        mutate(cv_expr = round(cv_expr, 3),
               mean_expr = round(mean_expr, 1)) %>%
        arrange(species, desc(cv_expr)), 
      caption = "Top 20 Most Variable Calcification Genes by Species") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "400px")
```

## Heatmap of CV for Common Orthologs

```{r cv-heatmap, fig.height=12, fig.width=8}
# Create CV matrix for common orthologs
cv_matrix <- common_cv %>%
  select(group_id, species, cv_expr) %>%
  pivot_wider(names_from = species, values_from = cv_expr) %>%
  column_to_rownames("group_id") %>%
  drop_na()

# Get annotations
cv_annot <- ortho_annot %>%
  filter(group_id %in% rownames(cv_matrix)) %>%
  select(group_id, protein_name) %>%
  distinct() %>%
  mutate(protein_name = str_trunc(protein_name, 50))

# Create heatmap
if(nrow(cv_matrix) > 5) {
  pheatmap(
    as.matrix(cv_matrix),
    scale = "none",
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    color = viridis(100),
    main = "Expression CV for Common Calcification Orthologs",
    fontsize_row = 6,
    fontsize_col = 10,
    show_rownames = nrow(cv_matrix) <= 50
  )
}
```

# Summary and Conclusions

## Summary Statistics Table

```{r final-summary}
# Gene-level summary
gene_summary_final <- all_gene_cv %>%
  group_by(species) %>%
  summarise(
    `N Genes` = n(),
    `Mean CV` = round(mean(cv_expr, na.rm = TRUE), 3),
    `Median CV` = round(median(cv_expr, na.rm = TRUE), 3),
    `SD CV` = round(sd(cv_expr, na.rm = TRUE), 3),
    Level = "Gene"
  )

# Exon-level summary
exon_summary_final <- all_exon_cv %>%
  group_by(species) %>%
  summarise(
    `N Genes` = n_distinct(group_id, na.rm = TRUE),
    `Mean CV` = round(mean(cv_expr, na.rm = TRUE), 3),
    `Median CV` = round(median(cv_expr, na.rm = TRUE), 3),
    `SD CV` = round(sd(cv_expr, na.rm = TRUE), 3),
    Level = "Exon"
  )

final_summary <- bind_rows(gene_summary_final, exon_summary_final) %>%
  arrange(Level, species)

kable(final_summary, caption = "Summary of Expression Variability Analysis") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Hypothesis Evaluation

```{r hypothesis-evaluation, results='asis'}
# Calculate effect sizes (median CV ratios)
acro_median <- median(all_gene_cv$cv_expr[all_gene_cv$species == "Acropora"], na.rm = TRUE)
peve_median <- median(all_gene_cv$cv_expr[all_gene_cv$species == "Porites"], na.rm = TRUE)
pocillo_median <- median(all_gene_cv$cv_expr[all_gene_cv$species == "Pocillopora"], na.rm = TRUE)

cat("## Hypothesis Test Results\n\n")
cat("**Hypothesis:** Gene expression of calcification genes is more variable in *Acropora* and *Pocillopora* than *Porites*\n\n")

cat("### Gene-Level Results\n\n")
cat("- Acropora median CV:", round(acro_median, 3), "\n")
cat("- Porites median CV:", round(peve_median, 3), "\n")
cat("- Pocillopora median CV:", round(pocillo_median, 3), "\n\n")

cat("- Acropora/Porites CV ratio:", round(acro_median/peve_median, 2), "\n")
cat("- Pocillopora/Porites CV ratio:", round(pocillo_median/peve_median, 2), "\n\n")

# Interpretation
cat("### Interpretation\n\n")
if(acro_median > peve_median & pocillo_median > peve_median) {
  cat("**Support for hypothesis:** Both *Acropora* and *Pocillopora* show higher expression variability than *Porites* in calcification genes.\n\n")
} else if(acro_median > peve_median | pocillo_median > peve_median) {
  cat("**Partial support for hypothesis:** Only one of *Acropora* or *Pocillopora* shows higher expression variability than *Porites*.\n\n")
} else {
  cat("**No support for hypothesis:** Neither *Acropora* nor *Pocillopora* shows higher expression variability than *Porites*.\n\n")
}

cat("### Exon-Level Results\n\n")
acro_exon_median <- median(all_exon_cv$cv_expr[all_exon_cv$species == "Acropora"], na.rm = TRUE)
peve_exon_median <- median(all_exon_cv$cv_expr[all_exon_cv$species == "Porites"], na.rm = TRUE)
pocillo_exon_median <- median(all_exon_cv$cv_expr[all_exon_cv$species == "Pocillopora"], na.rm = TRUE)

cat("- Acropora median exon CV:", round(acro_exon_median, 3), "\n")
cat("- Porites median exon CV:", round(peve_exon_median, 3), "\n")
cat("- Pocillopora median exon CV:", round(pocillo_exon_median, 3), "\n\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
