---
title: "57-Hyp-cal-gbm"
subtitle: "Testing hypothesis: Porites exhibits constitutive higher GBM in calcification genes"
author: "Steven Roberts"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

# Introduction

## Hypothesis

**Porites (peve) exhibits constitutive higher gene body methylation (GBM) in key calcification genes (e.g., Ca-ATPase, Carbonic Anhydrase) than Acropora (apul) and Pocillopora (ptua) compared to other genes.**

This analysis will:

1. Load gene body methylation (GBM) data for all three coral species
2. Identify calcification-related genes from the biomineralization gene lists
3. Compare GBM levels between calcification genes and all other genes across species
4. Test whether Porites shows constitutively higher GBM in calcification genes relative to other species

## Load Libraries

```{r load-libraries}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(viridis)
library(knitr)
library(kableExtra)
library(scales)
```

# Data Loading

## Load Biomineralization Gene Lists

These contain the calcification-related genes for each species.

```{r load-biomin}
# Load biomineralization gene counts for each species
apul_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/apul_biomin_counts.csv")
peve_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/peve_biomin_counts.csv")
ptua_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/ptua_biomin_counts.csv")

# Display summary
cat("Apul biomin genes:", nrow(apul_biomin), "\n")
cat("Peve biomin genes:", nrow(peve_biomin), "\n")
cat("Ptua biomin genes:", nrow(ptua_biomin), "\n")
```

## Load Gene Body Methylation Data

```{r load-gbm}
# Load GBM data for each species
apul_gbm <- read_tsv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/D-Apul/output/40-Apul-Gene-Methylation/Apul-gene-methylation_75pct.tsv")
peve_gbm <- read_tsv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/E-Peve/output/15-Peve-Gene-Methylation/Peve-gene-methylation_75pct.tsv")
ptua_gbm <- read_tsv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/F-Ptua/output/09-Ptua-Gene-Methylation/Ptua-gene-methylation_75pct.tsv")

# Display dimensions
cat("Apul GBM genes:", nrow(apul_gbm), "samples:", ncol(apul_gbm) - 1, "\n")
cat("Peve GBM genes:", nrow(peve_gbm), "samples:", ncol(peve_gbm) - 1, "\n")
cat("Ptua GBM genes:", nrow(ptua_gbm), "samples:", ncol(ptua_gbm) - 1, "\n")
```

## Load Ortholog Annotation

```{r load-annotation}
# Load ortholog annotation for cross-species gene information
ortho_annot <- read_csv("../output/12-ortho-annot/ortholog_groups_annotated.csv")

# Preview annotation structure
head(ortho_annot) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

# Data Processing

## Calculate Mean GBM per Gene

For each species, calculate the mean GBM across all samples for each gene.

```{r calc-mean-gbm}
# Function to calculate mean GBM per gene
calc_mean_gbm <- function(gbm_df) {
  gene_col <- names(gbm_df)[1]  # First column is gene ID
  sample_cols <- names(gbm_df)[-1]  # Remaining columns are samples
  
  gbm_df %>%
    rowwise() %>%
    mutate(mean_gbm = mean(c_across(all_of(sample_cols)), na.rm = TRUE)) %>%
    ungroup() %>%
    select(gene_id = all_of(gene_col), mean_gbm)
}

# Calculate mean GBM for each species
apul_mean_gbm <- calc_mean_gbm(apul_gbm)
peve_mean_gbm <- calc_mean_gbm(peve_gbm)
ptua_mean_gbm <- calc_mean_gbm(ptua_gbm)

# Summary statistics
cat("Apul mean GBM summary:\n")
summary(apul_mean_gbm$mean_gbm)
cat("\nPeve mean GBM summary:\n")
summary(peve_mean_gbm$mean_gbm)
cat("\nPtua mean GBM summary:\n")
summary(ptua_mean_gbm$mean_gbm)
```

## Identify Calcification Genes

Extract gene IDs from biomineralization lists and classify genes as calcification or non-calcification.

```{r identify-calc-genes}
# Extract biomin gene IDs for each species
apul_biomin_genes <- apul_biomin[[1]]  # First column contains gene IDs
peve_biomin_genes <- peve_biomin[[1]]
ptua_biomin_genes <- ptua_biomin[[1]]

# Add gene category to GBM data
apul_mean_gbm <- apul_mean_gbm %>%
  mutate(
    gene_category = ifelse(gene_id %in% apul_biomin_genes, "Calcification", "Other"),
    species = "Apul"
  )

peve_mean_gbm <- peve_mean_gbm %>%
  mutate(
    gene_category = ifelse(gene_id %in% peve_biomin_genes, "Calcification", "Other"),
    species = "Peve"
  )

ptua_mean_gbm <- ptua_mean_gbm %>%
  mutate(
    gene_category = ifelse(gene_id %in% ptua_biomin_genes, "Calcification", "Other"),
    species = "Ptua"
  )

# Count genes in each category
cat("Gene counts by category:\n")
cat("Apul - Calcification:", sum(apul_mean_gbm$gene_category == "Calcification"), 
    "Other:", sum(apul_mean_gbm$gene_category == "Other"), "\n")
cat("Peve - Calcification:", sum(peve_mean_gbm$gene_category == "Calcification"), 
    "Other:", sum(peve_mean_gbm$gene_category == "Other"), "\n")
cat("Ptua - Calcification:", sum(ptua_mean_gbm$gene_category == "Calcification"), 
    "Other:", sum(ptua_mean_gbm$gene_category == "Other"), "\n")
```

## Combine Data

```{r combine-data}
# Combine all species data
combined_gbm <- bind_rows(apul_mean_gbm, peve_mean_gbm, ptua_mean_gbm)

# Set factor levels for consistent ordering
combined_gbm$species <- factor(combined_gbm$species, levels = c("Apul", "Peve", "Ptua"))
combined_gbm$gene_category <- factor(combined_gbm$gene_category, levels = c("Calcification", "Other"))

# Remove NA values
combined_gbm <- combined_gbm %>% filter(!is.na(mean_gbm))

# Summary
cat("Combined dataset dimensions:", nrow(combined_gbm), "genes\n")
```

# Summary Statistics

## GBM Summary by Species and Gene Category

```{r summary-stats}
# Calculate summary statistics
gbm_summary <- combined_gbm %>%
  group_by(species, gene_category) %>%
  summarise(
    n = n(),
    mean_gbm = mean(mean_gbm, na.rm = TRUE),
    sd_gbm = sd(mean_gbm, na.rm = TRUE),
    median_gbm = median(mean_gbm, na.rm = TRUE),
    q25 = quantile(mean_gbm, 0.25, na.rm = TRUE),
    q75 = quantile(mean_gbm, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

gbm_summary %>%
  kable(digits = 2, caption = "GBM Summary Statistics by Species and Gene Category") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## GBM Ratio: Calcification vs Other Genes

```{r gbm-ratio}
# Calculate ratio of calcification GBM to other genes GBM for each species
gbm_ratio <- gbm_summary %>%
  select(species, gene_category, mean_gbm) %>%
  pivot_wider(names_from = gene_category, values_from = mean_gbm)

# Check column names and use backticks for column references
gbm_ratio <- gbm_ratio %>%
  mutate(
    ratio_calc_to_other = `Calcification` / `Other`,
    diff_calc_minus_other = `Calcification` - `Other`
  )

gbm_ratio %>%
  kable(digits = 3, caption = "GBM Ratio: Calcification Genes vs Other Genes") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

# Visualizations

## Distribution of GBM by Species and Gene Category

```{r violin-plot, fig.width=12, fig.height=8}
# Violin plot comparing GBM distributions
ggplot(combined_gbm, aes(x = species, y = mean_gbm, fill = gene_category)) +
  geom_violin(trim = FALSE, alpha = 0.7, position = position_dodge(width = 0.8)) +
  geom_boxplot(width = 0.15, position = position_dodge(width = 0.8), 
               outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = c("Calcification" = "#E74C3C", "Other" = "#3498DB")) +
  labs(
    title = "Gene Body Methylation Distribution by Species and Gene Category",
    subtitle = "Comparing calcification genes vs other genes across coral species",
    x = "Species",
    y = "Mean Gene Body Methylation (%)",
    fill = "Gene Category"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 14),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

## Boxplot with Statistical Comparisons

```{r boxplot-stats, fig.width=12, fig.height=8}
# Boxplot with statistical comparisons
ggplot(combined_gbm, aes(x = interaction(species, gene_category), y = mean_gbm, fill = gene_category)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("Calcification" = "#E74C3C", "Other" = "#3498DB")) +
  scale_x_discrete(labels = c("Apul\nCalc", "Apul\nOther", "Peve\nCalc", "Peve\nOther", "Ptua\nCalc", "Ptua\nOther")) +
  labs(
    title = "Gene Body Methylation: Calcification vs Other Genes",
    x = "Species and Gene Category",
    y = "Mean Gene Body Methylation (%)",
    fill = "Gene Category"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 14),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

## Species Comparison - Calcification Genes Only

```{r calc-genes-comparison, fig.width=10, fig.height=6}
# Filter to calcification genes only
calc_genes_only <- combined_gbm %>% filter(gene_category == "Calcification")

# Violin plot for calcification genes across species
ggplot(calc_genes_only, aes(x = species, y = mean_gbm, fill = species)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.3, size = 1) +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "Gene Body Methylation in Calcification Genes",
    subtitle = "Comparing GBM levels across coral species",
    x = "Species",
    y = "Mean Gene Body Methylation (%)"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 14),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

## Density Plot Comparison

```{r density-plot, fig.width=12, fig.height=8}
# Density plot by species and category
ggplot(combined_gbm, aes(x = mean_gbm, fill = gene_category)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~species, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c("Calcification" = "#E74C3C", "Other" = "#3498DB")) +
  labs(
    title = "Density Distribution of Gene Body Methylation",
    x = "Mean Gene Body Methylation (%)",
    y = "Density",
    fill = "Gene Category"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 14),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(face = "bold", size = 12)
  )
```

# Statistical Analysis

## Within-Species Comparison: Calcification vs Other Genes

Test whether calcification genes have significantly different GBM than other genes within each species.

```{r within-species-test}
# Wilcoxon rank-sum test for each species
within_species_tests <- combined_gbm %>%
  group_by(species) %>%
  summarise(
    W_statistic = wilcox.test(mean_gbm[gene_category == "Calcification"], 
                              mean_gbm[gene_category == "Other"])$statistic,
    p_value = wilcox.test(mean_gbm[gene_category == "Calcification"], 
                          mean_gbm[gene_category == "Other"])$p.value,
    calc_median = median(mean_gbm[gene_category == "Calcification"], na.rm = TRUE),
    other_median = median(mean_gbm[gene_category == "Other"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    p_adjusted = p.adjust(p_value, method = "BH"),
    significance = case_when(
      p_adjusted < 0.001 ~ "***",
      p_adjusted < 0.01 ~ "**",
      p_adjusted < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

within_species_tests %>%
  kable(digits = 4, caption = "Within-Species Comparison: Calcification vs Other Genes (Wilcoxon Test)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Between-Species Comparison: Calcification Genes

Test whether GBM in calcification genes differs significantly between species.

```{r between-species-calc}
# Kruskal-Wallis test for calcification genes across species
calc_kruskal <- kruskal.test(mean_gbm ~ species, data = calc_genes_only)

cat("Kruskal-Wallis Test for Calcification Genes across Species:\n")
cat("Chi-squared:", calc_kruskal$statistic, "\n")
cat("df:", calc_kruskal$parameter, "\n")
cat("p-value:", calc_kruskal$p.value, "\n")

# Pairwise Wilcoxon tests for calcification genes
calc_pairwise <- pairwise.wilcox.test(calc_genes_only$mean_gbm, calc_genes_only$species, 
                                       p.adjust.method = "BH")

cat("\nPairwise Wilcoxon Tests (BH-adjusted p-values):\n")
print(calc_pairwise)
```

## Between-Species Comparison: All Genes

```{r between-species-all}
# Kruskal-Wallis test for all genes across species
all_kruskal <- kruskal.test(mean_gbm ~ species, data = combined_gbm)

cat("Kruskal-Wallis Test for All Genes across Species:\n")
cat("Chi-squared:", all_kruskal$statistic, "\n")
cat("df:", all_kruskal$parameter, "\n")
cat("p-value:", all_kruskal$p.value, "\n")

# Pairwise Wilcoxon tests for all genes
all_pairwise <- pairwise.wilcox.test(combined_gbm$mean_gbm, combined_gbm$species, 
                                      p.adjust.method = "BH")

cat("\nPairwise Wilcoxon Tests (BH-adjusted p-values):\n")
print(all_pairwise)
```

## Interaction Analysis: Species x Gene Category

```{r interaction-analysis}
# Test for interaction effect using aligned rank transform or similar
# First, let's look at the effect sizes

# Calculate difference in GBM (Calc - Other) for each species
diff_by_species <- combined_gbm %>%
  group_by(species, gene_category) %>%
  summarise(mean_gbm = mean(mean_gbm, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = gene_category, values_from = mean_gbm) %>%
  mutate(difference = `Calcification` - `Other`)

cat("Difference in Mean GBM (Calcification - Other) by Species:\n")
diff_by_species %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

# Hypothesis Evaluation

## Summary Plot: GBM Enrichment in Calcification Genes

```{r enrichment-plot, fig.width=10, fig.height=6}
# Calculate fold enrichment
enrichment_data <- gbm_summary %>%
  select(species, gene_category, mean_gbm) %>%
  pivot_wider(names_from = gene_category, values_from = mean_gbm) %>%
  mutate(
    fold_enrichment = `Calcification` / `Other`,
    percent_increase = (`Calcification` - `Other`) / `Other` * 100
  )

# Plot fold enrichment
ggplot(enrichment_data, aes(x = species, y = fold_enrichment, fill = species)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "GBM Enrichment in Calcification Genes",
    subtitle = "Ratio of mean GBM in calcification genes vs other genes",
    x = "Species",
    y = "Fold Enrichment (Calcification / Other)"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 14),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

## Comparison Summary

```{r comparison-summary}
# Create comprehensive comparison table
comparison_summary <- gbm_summary %>%
  select(species, gene_category, n, mean_gbm, median_gbm) %>%
  arrange(species, gene_category)

comparison_summary %>%
  kable(digits = 2, caption = "Comprehensive GBM Comparison Summary",
        col.names = c("Species", "Gene Category", "N Genes", "Mean GBM (%)", "Median GBM (%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  pack_rows("Acropora pulchra", 1, 2) %>%
  pack_rows("Porites evermanni", 3, 4) %>%
  pack_rows("Pocillopora tuahiniensis", 5, 6)
```

# Conclusions

```{r conclusions}
# Generate automated conclusions based on results
cat("=" , rep("=", 70), "=\n", sep = "")
cat("HYPOTHESIS EVALUATION\n")
cat("=" , rep("=", 70), "=\n\n", sep = "")

cat("Hypothesis: Porites (Peve) exhibits constitutive higher gene body methylation\n")
cat("in key calcification genes than Acropora (Apul) and Pocillopora (Ptua)\n")
cat("compared to other genes.\n\n")

cat("-", rep("-", 70), "-\n", sep = "")
cat("RESULTS SUMMARY\n")
cat("-", rep("-", 70), "-\n\n", sep = "")

# Get Peve calcification gene GBM
peve_calc_gbm <- gbm_summary %>% 
  filter(species == "Peve", gene_category == "Calcification") %>% 
  pull(mean_gbm)
peve_other_gbm <- gbm_summary %>% 
  filter(species == "Peve", gene_category == "Other") %>% 
  pull(mean_gbm)

apul_calc_gbm <- gbm_summary %>% 
  filter(species == "Apul", gene_category == "Calcification") %>% 
  pull(mean_gbm)
ptua_calc_gbm <- gbm_summary %>% 
  filter(species == "Ptua", gene_category == "Calcification") %>% 
  pull(mean_gbm)

cat("1. Mean GBM in Calcification Genes:\n")
cat("   - Apul:", round(apul_calc_gbm, 2), "%\n")
cat("   - Peve:", round(peve_calc_gbm, 2), "%\n")
cat("   - Ptua:", round(ptua_calc_gbm, 2), "%\n\n")

cat("2. GBM Enrichment (Calcification/Other ratio):\n")
for (i in 1:nrow(enrichment_data)) {
  cat("   -", enrichment_data$species[i], ":", round(enrichment_data$fold_enrichment[i], 3), "\n")
}

cat("\n3. Statistical Significance:\n")
cat("   - Kruskal-Wallis test (calcification genes across species): p =", 
    format(calc_kruskal$p.value, scientific = TRUE), "\n")

cat("\n", "-", rep("-", 70), "-\n", sep = "")
cat("CONCLUSION\n")
cat("-", rep("-", 70), "-\n", sep = "")

# Determine if hypothesis is supported
if (peve_calc_gbm > apul_calc_gbm && peve_calc_gbm > ptua_calc_gbm) {
  cat("\nThe data SUPPORT the hypothesis that Porites (Peve) has higher GBM in\n")
  cat("calcification genes compared to other species.\n")
} else {
  cat("\nThe data DO NOT SUPPORT the hypothesis that Porites (Peve) has higher GBM\n")
  cat("in calcification genes compared to other species.\n")
}
```

# Session Info

```{r session-info}
sessionInfo()
```
