---
title: "58-miRNA expression"
author: "Jill Ashey"
date: "2026-01-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

# Introduction 

This analysis will test the following hypothese: 

1. Acropora will have higher miRNA expression relative to Porites and Pocillopora 
2. Acropora will have more variable miRNA expression relative to Porites and Pocillopora

# Load libraries 

```{r load-libraries}
library(tidyverse)
library(stringr)
library(DESeq2)
library(kableExtra)
```

# Data preprocessing 

Read in data and subset miRNAs 
```{r load-miRNA-counts}
# Read in raw counts
apul_mirna_results <- read_delim("../../D-Apul/output/04-Apul-sRNA-discovery-ShortStack/ShortStack_out/Counts.txt", delim="\t") 
peve_mirna_results <- read_delim("../../E-Peve/output/08-Peve-sRNA-discovery-ShortStack/ShortStack_out/Counts.txt", delim="\t") 
ptua_mirna_results <- read_delim("../../F-Ptua/output/04-Ptua-sRNA-discovery-ShortStack/ShortStack_out/Counts.txt", delim="\t") 

# Subset rows where MIRNA=Y and remove unneeded cols
apul_mirna_results <- apul_mirna_results %>%
  filter(MIRNA == "Y") %>%
  select(-c(Coords, MIRNA))
peve_mirna_results <- peve_mirna_results %>%
  filter(MIRNA == "Y") %>%
  select(-c(Coords, MIRNA))
ptua_mirna_results <- ptua_mirna_results %>%
  filter(MIRNA == "Y") %>%
  select(-c(Coords, MIRNA))

# Remove unneeded info from column name
colnames(apul_mirna_results) <- sub("-fastp-adapters-polyG-31bp-merged_condensed", "", colnames(apul_mirna_results))
colnames(peve_mirna_results) <- sub("-fastp-adapters-polyG-31bp-merged_condensed", "", colnames(peve_mirna_results))
colnames(ptua_mirna_results) <- sub("-fastp-adapters-polyG-31bp-merged_condensed", "", colnames(ptua_mirna_results))

# Check out data 
head(apul_mirna_results)
head(peve_mirna_results)
head(ptua_mirna_results)
```

Use column names to rename columns based on RNA metadatta 
```{r rename-columns}
rna_meta <- read.csv("../data/rna_metadata.csv", header = T) %>%
  mutate(Colony_TP = paste(ColonyID, Timepoint, sep = "_")) %>%
  select(AzentaSampleName, Colony_TP)

# Make vector of names 
sample_lookup <- setNames(
  rna_meta$Colony_TP,
  rna_meta$AzentaSampleName)

# Rename columns 
apul_mirna_results_renamed <- apul_mirna_results %>%
  rename_with(
    ~ ifelse(.x %in% names(sample_lookup), sample_lookup[.x], .x), -Name)
peve_mirna_results_renamed <- peve_mirna_results %>%
  rename_with(
    ~ ifelse(.x %in% names(sample_lookup), sample_lookup[.x], .x), -Name)
ptua_mirna_results_renamed <- ptua_mirna_results %>%
  rename_with(
    ~ ifelse(.x %in% names(sample_lookup), sample_lookup[.x], .x), -Name)

head(apul_mirna_results_renamed)
head(peve_mirna_results_renamed)
head(ptua_mirna_results_renamed)
```

# Calculate CV 

Function to calculate CV
```{r cv-function}
cv <- function(x) {
  x <- x[!is.na(x)]
  if (mean(x) == 0) return(NA_real_)
  sd(x) / mean(x)
}
```

Pivot df to long format
```{r pivot}
apul_mirna_long <- apul_mirna_results_renamed %>%
  pivot_longer(
    -Name,
    names_to = "Sample",
    values_to = "count") %>%
  mutate(TP = str_extract(Sample, "TP[1-4]"))

peve_mirna_long <- peve_mirna_results_renamed %>%
  pivot_longer(
    -Name,
    names_to = "Sample",
    values_to = "count") %>%
  mutate(TP = str_extract(Sample, "TP[1-4]"))

ptua_mirna_long <- ptua_mirna_results_renamed %>%
  pivot_longer(
    -Name,
    names_to = "Sample",
    values_to = "count") %>%
  mutate(TP = str_extract(Sample, "TP[1-4]"))
```

## Calculate mean, sd, and CV across all samples 

```{r mean-sd-cv-summary}
calc_cv_by_tp_normalized <- function(count_df, species_name,
                                     min_mean = 5) {

  ## ---------------------------
  ## Prepare count matrix
  ## ---------------------------
  count_mat <- count_df %>%
    column_to_rownames("Name") %>%
    as.matrix()

  ## Remove rows that are all zero
  count_mat <- count_mat[rowSums(count_mat) > 0, , drop = FALSE]

  ## Remove columns (samples) that are all zero
  count_mat <- count_mat[, colSums(count_mat) > 0, drop = FALSE]

  ## Safety check
  if (ncol(count_mat) < 2) {
    stop(paste("Not enough non-zero samples for", species_name))
  }

  coldata <- data.frame(
    sample = colnames(count_mat),
    row.names = colnames(count_mat)
  )

  ## ---------------------------
  ## DESeq2 normalization
  ## ---------------------------
  dds <- DESeqDataSetFromMatrix(
    countData = round(count_mat),
    colData   = coldata,
    design    = ~ 1
  )

  dds <- estimateSizeFactors(dds, type = "poscounts")

  ## Check for NA size factors (should now be gone)
  if (any(is.na(sizeFactors(dds)))) {
    stop(paste("NA size factors remain for", species_name))
  }

  norm_counts <- counts(dds, normalized = TRUE)

  ## ---------------------------
  ## Filter low-expression clusters
  ## ---------------------------
  keep <- rowMeans(norm_counts) >= min_mean
  norm_counts <- norm_counts[keep, , drop = FALSE]

  ## ---------------------------
  ## Long format + CV by TP
  ## ---------------------------
  norm_counts %>%
    as.data.frame() %>%
    rownames_to_column("Name") %>%
    pivot_longer(
      cols = -Name,
      names_to = "Sample",
      values_to = "norm_count"
    ) %>%
    mutate(
      TP = str_extract(Sample, "TP[1-4]"),
      Species = species_name
    ) %>%
    group_by(Name, TP, Species) %>%
    summarise(
      mean_expr = mean(norm_count, na.rm = TRUE),
      sd_expr   = sd(norm_count, na.rm = TRUE),
      cv_expr   = sd_expr / mean_expr,
      n_samples = n(),
      .groups = "drop"
    )
}


# Run for all spp
cv_all_species <- bind_rows(
  calc_cv_by_tp_normalized(apul_mirna_results_renamed, "Acropora"),
  calc_cv_by_tp_normalized(peve_mirna_results_renamed, "Porites"),
  calc_cv_by_tp_normalized(ptua_mirna_results_renamed, "Pocillopora"))
```

Look at CV across all spp, regardless of timepoint 
```{r cv-all-samples}
cv_summary <- cv_all_species %>%
  group_by(Species) %>%
  summarise(
    n_mirna = n_distinct(Name),  
    mean_cv = mean(cv_expr, na.rm = TRUE),
    median_cv = median(cv_expr, na.rm = TRUE),
    sd_cv = sd(cv_expr, na.rm = TRUE),
    iqr_cv = IQR(cv_expr, na.rm = TRUE),
    q25 = quantile(cv_expr, 0.25, na.rm = TRUE),
    q75 = quantile(cv_expr, 0.75, na.rm = TRUE))

kable(cv_summary, digits = 3, caption = "miRNA Expression CV Summary by Species") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Violin plot of CV distribution
p1 <- ggplot(cv_all_species, aes(x = Species, y = cv_expr, fill = Species)) +
  geom_violin(alpha = 0.7, trim = TRUE) +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
  scale_fill_manual(values = c("Acropora" = "#E64B35", 
                                "Porites" = "#4DBBD5", 
                                "Pocillopora" = "#00A087")) +
  labs(
    title = "miRNA Expression Variability (CV)",
    subtitle = "Higher CV indicates more variable expression across samples",
    x = "Species",
    y = "Coefficient of Variation (CV)"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

print(p1)
```

Look at CV across all spp, including timepoint
```{r cv-timepoints}
cv_tp_summary <- cv_all_species %>%
  group_by(Species, TP) %>%
  summarise(
    n_mirna = n_distinct(Name),  
    mean_cv = mean(cv_expr, na.rm = TRUE),
    median_cv = median(cv_expr, na.rm = TRUE),
    sd_cv = sd(cv_expr, na.rm = TRUE),
    iqr_cv = IQR(cv_expr, na.rm = TRUE),
    q25 = quantile(cv_expr, 0.25, na.rm = TRUE),
    q75 = quantile(cv_expr, 0.75, na.rm = TRUE))

kable(cv_tp_summary, digits = 3, caption = "miRNA Expression CV Summary by Species and Timepoint") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Violin plot of CV distribution
p2 <- ggplot(cv_all_species, aes(x = Species, y = cv_expr, fill = TP)) +
  geom_violin(alpha = 0.7, trim = TRUE) +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
  #scale_fill_manual(values = c("Acropora" = "#E64B35", 
  #                              "Porites" = "#4DBBD5", 
  #                              "Pocillopora" = "#00A087")) +
  labs(
    title = "miRNA Expression Variability (CV)",
    subtitle = "Higher CV indicates more variable expression across timepoints",
    x = "Species",
    y = "Coefficient of Variation (CV)"
  ) +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

print(p2)
```

# Statistical tests for CV differences in miRNAs

Kruskal-Wallis test, regardless of timepoint 
```{r stats-all}
# Kruskal-Wallis tests
kw_test <- kruskal.test(cv_expr ~ Species, data = cv_all_species)
cat("Kruskal-Wallis test for CV differences:\n")
print(kw_test)

# Compare specific pairs
acro_vs_porites <- wilcox.test(
  cv_all_species$cv_expr[cv_all_species$Species == "Acropora"],
  cv_all_species$cv_expr[cv_all_species$Species == "Porites"]
)

acro_vs_poc <- wilcox.test(
  cv_all_species$cv_expr[cv_all_species$Species == "Acropora"],
  cv_all_species$cv_expr[cv_all_species$Species == "Pocillopora"]
)

pocillo_vs_porites <- wilcox.test(
  cv_all_species$cv_expr[cv_all_species$Species == "Pocillopora"],
  cv_all_species$cv_expr[cv_all_species$Species == "Porites"]
)

cat("\nAcropora vs Porites p-value:", acro_vs_porites$p.value, "\n")
cat("\nAcropora vs Pocillopora p-value:", acro_vs_poc$p.value, "\n")
cat("\nPocillopora vs Porites p-value:", pocillo_vs_porites$p.value, "\n")

```

Kruskal-Wallis test, including timepoint
```{r stats-tp}
# Get all unique timepoints and species
timepoints <- sort(unique(cv_all_species$TP))
species <- sort(unique(cv_all_species$Species))

# Initialize empty list to store results
kw_results <- list()
pairwise_results <- list()

# Loop through each timepoint
for(tp in timepoints){
  
  df_tp <- cv_all_species %>% filter(TP == tp)
  
  cat("=== Timepoint:", tp, "===\n")
  
  # Kruskal-Wallis test
  kw <- kruskal.test(cv_expr ~ Species, data = df_tp)
  print(kw)
  
  kw_results[[tp]] <- data.frame(
    TP = tp,
    KW_statistic = kw$statistic,
    KW_p = kw$p.value
  )
  
  # Pairwise Wilcoxon tests for all species pairs
  pairs <- combn(species, 2, simplify = FALSE)
  
  pairwise_tp <- data.frame(
    TP = character(),
    Species1 = character(),
    Species2 = character(),
    p_value = numeric(),
    stringsAsFactors = FALSE
  )
  
  for(pair in pairs){
    sp1 <- pair[1]
    sp2 <- pair[2]
    
    w <- wilcox.test(
      cv_expr ~ Species,
      data = df_tp %>% filter(Species %in% c(sp1, sp2))
    )
    
    pairwise_tp <- rbind(pairwise_tp,
                         data.frame(
                           TP = tp,
                           Species1 = sp1,
                           Species2 = sp2,
                           p_value = w$p.value
                         ))
  }
  
  pairwise_results[[tp]] <- pairwise_tp
  cat("\n")
}

# Combine results into tidy tables
kw_results_df <- bind_rows(kw_results)
pairwise_results_df <- bind_rows(pairwise_results)

# Uses p < 0.05 threshold
pairwise_results_df <- pairwise_results_df %>%
  mutate(signif = ifelse(p_value < 0.05, "*", "ns"))

# View summary tables
kw_results_df
pairwise_results_df

```

# Hypothesis interpretation 

Summary, regardless of timepoint 
```{r hypothesis-testing}
cv_medians <- cv_all_species %>%
  group_by(Species) %>%
  summarise(median_cv = median(cv_expr, na.rm = TRUE)) %>%
  pivot_wider(names_from = Species, values_from = median_cv)

# Extract median values
acro_median   <- cv_medians$Acropora
peve_median   <- cv_medians$Porites
pocillo_median <- cv_medians$Pocillopora

# Test hypothesis
if(acro_median > peve_median & pocillo_median > peve_median) {
  cat("**Support for hypothesis:** Both *Acropora* and *Pocillopora* show higher miRNA expression variability than *Porites*.\n")
} else if(acro_median > peve_median | pocillo_median > peve_median) {
  cat("**Partial support for hypothesis:** Only one of *Acropora* or *Pocillopora* shows higher miRNA expression variability than *Porites*.\n")
} else {
  cat("**No support for hypothesis:** Neither *Acropora* nor *Pocillopora* shows higher miRNA expression variability than *Porites*.\n")
}
```

Summary including timepoint 
```{r hypothesis-testing-again}
for(tp in unique(cv_all_species$TP)) {
  
  cv_tp <- cv_all_species %>% filter(TP == tp)
  
  cv_medians <- cv_tp %>%
    group_by(Species) %>%
    summarise(median_cv = median(cv_expr, na.rm = TRUE)) %>%
    pivot_wider(names_from = Species, values_from = median_cv)
  
  acro_median   <- cv_medians$Acropora
  peve_median   <- cv_medians$Porites
  pocillo_median <- cv_medians$Pocillopora
  
  cat("\nTimepoint:", tp, "\n")
  
  if(acro_median > peve_median & pocillo_median > peve_median) {
    cat("**Support for hypothesis:** Both *Acropora* and *Pocillopora* show higher miRNA expression variability than *Porites*.\n")
  } else if(acro_median > peve_median | pocillo_median > peve_median) {
    cat("**Partial support for hypothesis:** Only one of *Acropora* or *Pocillopora* shows higher miRNA expression variability than *Porites*.\n")
  } else {
    cat("**No support for hypothesis:** Neither *Acropora* nor *Pocillopora* shows higher miRNA expression variability than *Porites*.\n")
  }
}
```

# Session info 

```{r session}
sessionInfo()
```

