---
title: "58.1-expression of miRNAs interacting with biomin genes"
author: "Jill Ashey"
date: "2026-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

# Introduction 

This analysis will test the following hypothesis: 

- Expression of miRNAs that putatively interact with biomineralization genes will be more variable in Acropora than in Porites or Pocillopora

# Load libraries 

```{r load-libraries}
library(tidyverse)
library(stringr)
library(DESeq2)
library(kableExtra)
library(purrr)
library(readr)
```

# Data preprocessing - miRNA counts data 

Read in data and subset miRNAs 
```{r load-miRNA-counts}
# Read in raw counts
apul_mirna_results <- read_delim("../../D-Apul/output/04-Apul-sRNA-discovery-ShortStack/ShortStack_out/Counts.txt", delim="\t") 
peve_mirna_results <- read_delim("../../E-Peve/output/08-Peve-sRNA-discovery-ShortStack/ShortStack_out/Counts.txt", delim="\t") 
ptua_mirna_results <- read_delim("../../F-Ptua/output/04-Ptua-sRNA-discovery-ShortStack/ShortStack_out/Counts.txt", delim="\t") 

# Subset rows where MIRNA=Y and remove unneeded cols
apul_mirna_results <- apul_mirna_results %>%
  filter(MIRNA == "Y") %>%
  select(-c(Coords, MIRNA))
peve_mirna_results <- peve_mirna_results %>%
  filter(MIRNA == "Y") %>%
  select(-c(Coords, MIRNA))
ptua_mirna_results <- ptua_mirna_results %>%
  filter(MIRNA == "Y") %>%
  select(-c(Coords, MIRNA))

# Remove unneeded info from column name
colnames(apul_mirna_results) <- sub("-fastp-adapters-polyG-31bp-merged_condensed", "", colnames(apul_mirna_results))
colnames(peve_mirna_results) <- sub("-fastp-adapters-polyG-31bp-merged_condensed", "", colnames(peve_mirna_results))
colnames(ptua_mirna_results) <- sub("-fastp-adapters-polyG-31bp-merged_condensed", "", colnames(ptua_mirna_results))

# Check out data 
head(apul_mirna_results)
head(peve_mirna_results)
head(ptua_mirna_results)
```

Use column names to rename columns based on RNA metadatta 
```{r rename-columns}
rna_meta <- read.csv("../data/rna_metadata.csv", header = T) %>%
  mutate(Colony_TP = paste(ColonyID, Timepoint, sep = "_")) %>%
  select(AzentaSampleName, Colony_TP)

# Make vector of names 
sample_lookup <- setNames(
  rna_meta$Colony_TP,
  rna_meta$AzentaSampleName)

# Rename columns 
apul_mirna_results_renamed <- apul_mirna_results %>%
  rename_with(
    ~ ifelse(.x %in% names(sample_lookup), sample_lookup[.x], .x), -Name)
peve_mirna_results_renamed <- peve_mirna_results %>%
  rename_with(
    ~ ifelse(.x %in% names(sample_lookup), sample_lookup[.x], .x), -Name)
ptua_mirna_results_renamed <- ptua_mirna_results %>%
  rename_with(
    ~ ifelse(.x %in% names(sample_lookup), sample_lookup[.x], .x), -Name)

head(apul_mirna_results_renamed)
head(peve_mirna_results_renamed)
head(ptua_mirna_results_renamed)
```

# miRNAs interacting with biomineralization genes 

Read in miranda data for all spp
```{r}
# Column names for miranda data files 
cols <- c(
  "miRNA", "mRNA", "score", "energy",
  "query_start_end", "subject_start_end",
  "total_bp_shared", "query_similar", "subject_similar")

# Apul miranda data 
apul_files <- tibble(
  path = c(
    "../../D-Apul/output/07-Apul-miRNA-mRNA-miRanda/Apul-miRanda-3UTR-strict-parsed-geneIDs.txt",
    "../../D-Apul/output/07.1-Apul-miRNA-mRNA-miRanda-additional_inputs/Apul-miRanda-mRNA_full-strict-parsed.txt",
    "../../D-Apul/output/07.1-Apul-miRNA-mRNA-miRanda-additional_inputs/Apul-miRanda-5UTR_1kb-strict-parsed.txt"),
  Type = c("3UTR", "CDS", "5UTR"))

# Format Apul mRNA and miRNA names  
apul_mir_interact <- apul_files %>%
  mutate(data = map(path, ~ read_tsv(.x, col_names = cols))) %>%
  unnest(data) %>%
  mutate(miRNA = gsub("^>|\\..*$", "", miRNA),
         mRNA  = gsub("::.*$", "", mRNA))
```

NEED TO ID THE GENES ASSOCIATED WITH THE 5UTR AND CDS MIRANDA FILES using the GFFs.

# Intersect interaction data with biomineralization gene lists 

Read in biomin genes
```{r}
# Load biomin count data from GitHub
apul_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/apul_biomin_counts.csv")
apul_biomin_genes <- apul_biomin$gene_id
peve_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/peve_biomin_counts.csv")
ptua_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/ptua_biomin_counts.csv")
```

Filter miranda data based on biomin genes 
```{r}
apul_mir_interact_biomin <- apul_mir_interact %>%
  filter(mRNA %in% apul_biomin_genes)
length(unique(apul_mir_interact_biomin$miRNA))
length(unique(apul_mir_interact_biomin$mRNA))
```

# Filter miRNA counts based on miRNAs interacting with biomin genes

```{r}
apul_mir_interact_biomin_list <- apul_mir_interact_biomin$miRNA

apul_mirna_results_biomin <- apul_mirna_results_renamed %>%
  filter(Name %in% apul_mir_interact_biomin_list)
```





