---
title: "58.1-expression of miRNAs interacting with biomin genes"
author: "Jill Ashey"
date: "2026-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

# Introduction 

This analysis will test the following hypothesis: 

- Expression of miRNAs that putatively interact with biomineralization genes will be more variable in Acropora than in Porites or Pocillopora

# Load libraries 

```{r load-libraries}
library(tidyverse)
library(stringr)
library(DESeq2)
library(kableExtra)
library(purrr)
library(readr)
library(igraph)
```

# Data preprocessing - miRNA counts data 

Read in data and subset miRNAs 
```{r load-miRNA-counts}
# Read in raw counts
apul_mirna_results <- read_delim("../../D-Apul/output/04-Apul-sRNA-discovery-ShortStack/ShortStack_out/Counts.txt", delim="\t") 
peve_mirna_results <- read_delim("../../E-Peve/output/08-Peve-sRNA-discovery-ShortStack/ShortStack_out/Counts.txt", delim="\t") 
ptua_mirna_results <- read_delim("../../F-Ptua/output/04-Ptua-sRNA-discovery-ShortStack/ShortStack_out/Counts.txt", delim="\t") 

# Subset rows where MIRNA=Y and remove unneeded cols
apul_mirna_results <- apul_mirna_results %>%
  filter(MIRNA == "Y") %>%
  select(-c(Coords, MIRNA))
peve_mirna_results <- peve_mirna_results %>%
  filter(MIRNA == "Y") %>%
  select(-c(Coords, MIRNA))
ptua_mirna_results <- ptua_mirna_results %>%
  filter(MIRNA == "Y") %>%
  select(-c(Coords, MIRNA))

# Remove unneeded info from column name
colnames(apul_mirna_results) <- sub("-fastp-adapters-polyG-31bp-merged_condensed", "", colnames(apul_mirna_results))
colnames(peve_mirna_results) <- sub("-fastp-adapters-polyG-31bp-merged_condensed", "", colnames(peve_mirna_results))
colnames(ptua_mirna_results) <- sub("-fastp-adapters-polyG-31bp-merged_condensed", "", colnames(ptua_mirna_results))

# Check out data 
head(apul_mirna_results)
head(peve_mirna_results)
head(ptua_mirna_results)
```

Use column names to rename columns based on RNA metadatta 
```{r rename-columns}
rna_meta <- read.csv("../data/rna_metadata.csv", header = T) %>%
  mutate(Colony_TP = paste(ColonyID, Timepoint, sep = "_")) %>%
  select(AzentaSampleName, Colony_TP)

# Make vector of names 
sample_lookup <- setNames(
  rna_meta$Colony_TP,
  rna_meta$AzentaSampleName)

# Rename columns 
apul_mirna_results_renamed <- apul_mirna_results %>%
  rename_with(
    ~ ifelse(.x %in% names(sample_lookup), sample_lookup[.x], .x), -Name)
peve_mirna_results_renamed <- peve_mirna_results %>%
  rename_with(
    ~ ifelse(.x %in% names(sample_lookup), sample_lookup[.x], .x), -Name)
ptua_mirna_results_renamed <- ptua_mirna_results %>%
  rename_with(
    ~ ifelse(.x %in% names(sample_lookup), sample_lookup[.x], .x), -Name)

head(apul_mirna_results_renamed)
head(peve_mirna_results_renamed)
head(ptua_mirna_results_renamed)
```

# miRNAs interacting with biomineralization genes 

Read in miranda data for all spp
```{r}
# Column names for miranda data files 
cols <- c(
  "miRNA", "mRNA", "score", "energy",
  "query_start_end", "subject_start_end",
  "total_bp_shared", "query_similar", "subject_similar")

# Apul miranda data 
apul_files <- tibble(
  path = c(
    "../../D-Apul/output/07-Apul-miRNA-mRNA-miRanda/Apul-miRanda-3UTR-strict-parsed-geneIDs.txt",
    "../../D-Apul/output/07.1-Apul-miRNA-mRNA-miRanda-additional_inputs/Apul-miRanda-mRNA_full-strict-parsed.txt",
    "../../D-Apul/output/07.1-Apul-miRNA-mRNA-miRanda-additional_inputs/Apul-miRanda-5UTR_1kb-strict-parsed.txt"),
  Type = c("3UTR", "CDS", "5UTR"))

# Format Apul mRNA and miRNA names  
apul_mir_interact <- apul_files %>%
  mutate(data = map(path, ~ read_tsv(.x, col_names = cols))) %>%
  unnest(data) %>%
  mutate(miRNA = gsub("^>|\\..*$", "", miRNA),
         mRNA  = gsub("::.*$", "", mRNA))
```

Read in associated gene ID information and merge with interaction information to get gene IDs. 
```{r}
apul_3utr_gene_id <- read.delim("../../D-Apul/output/05-Apul-annotate-UTRs/Apul-3UTR-FUNids.txt", header = F)[,1:4]
colnames(apul_3utr_gene_id) <- c("mRNA", "Type", "transcript_id", "gene_id")
apul_3utr_gene_id$gene_id <- gsub("Parent=", "", apul_3utr_gene_id$gene_id)
apul_3utr_gene_id$Type <- gsub("3prime_UTR", "3UTR", apul_3utr_gene_id$Type)


apul_5utr_gene_id <- read.delim("../../D-Apul/output/05-Apul-annotate-UTRs/Apul-5UTR-FUNids.txt", header = F)[,1:4]
colnames(apul_5utr_gene_id) <- c("mRNA", "Type", "transcript_id", "gene_id")
apul_5utr_gene_id$gene_id <- gsub("Parent=", "", apul_5utr_gene_id$gene_id)
apul_5utr_gene_id$Type <- gsub("5prime_UTR", "5UTR", apul_5utr_gene_id$Type)

apul_cds_gene_id <- read.delim("../../D-Apul/output/05-Apul-annotate-UTRs/Apul-mRNA-FUNids.txt", header = F)[,1:4]
colnames(apul_cds_gene_id) <- c("mRNA", "Type", "transcript_id", "gene_id")
apul_cds_gene_id$gene_id <- gsub("Parent=", "", apul_cds_gene_id$gene_id)
apul_cds_gene_id$Type <- gsub("mRNA", "CDS", apul_cds_gene_id$Type)

apul_gene_ids <- rbind(apul_3utr_gene_id, apul_5utr_gene_id, apul_cds_gene_id)

apul_mir_interact_gene_ids <- apul_mir_interact %>%
  inner_join(apul_gene_ids, by = c("mRNA", "Type"))
```

# Intersect interaction data with biomineralization gene lists 

Read in biomin genes
```{r}
# Load biomin count data from GitHub
apul_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/apul_biomin_counts.csv")
apul_biomin_genes <- apul_biomin$gene_id
peve_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/peve_biomin_counts.csv")
ptua_biomin <- read_csv("https://raw.githubusercontent.com/urol-e5/timeseries-molecular-calcification/refs/heads/main/M-multi-species/output/33-biomin-pathway-counts/ptua_biomin_counts.csv")
```

Filter miranda data based on biomin genes 
```{r}
apul_mir_interact_biomin <- apul_mir_interact_gene_ids %>%
  filter(gene_id %in% apul_biomin_genes) %>%
  filter(energy <= -20) %>%
  filter(query_similar >= 80)
length(unique(apul_mir_interact_biomin$miRNA))
length(unique(apul_mir_interact_biomin$mRNA))
```

48 miRNAs interact with 257 mRNAs. I filtered for 80% similarity of the query sequence, meaning that at least ~19 bp in the miRNA match the mRNA. This is almost perfect complementarity, which has been demonstrated in Nematostella. We can adjust this based on our interests. 

# Filter miRNA counts based on miRNAs interacting with biomin genes

```{r}
apul_mir_interact_biomin_list <- apul_mir_interact_biomin$miRNA

apul_mirna_results_biomin <- apul_mirna_results_renamed %>%
  filter(Name %in% apul_mir_interact_biomin_list)
```

# Calculate CV of remaining miRNAs

Function to calculate CV
```{r cv-function}
cv <- function(x) {
  x <- x[!is.na(x)]
  if (mean(x) == 0) return(NA_real_)
  sd(x) / mean(x)
}
```

Pivot df to long format
```{r pivot}
apul_mirna_results_biomin_long <- apul_mirna_results_biomin %>%
  pivot_longer(
    -Name,
    names_to = "Sample",
    values_to = "count") %>%
  mutate(TP = str_extract(Sample, "TP[1-4]"))
```

Calculate mean, sd, and CV across all samples 
```{r}
calc_cv_acropora <- function(long_df, min_mean = 5) {

  ## ---------------------------
  ## Back to count matrix
  ## ---------------------------
  count_mat <- long_df %>%
    select(Name, Sample, count) %>%
    pivot_wider(names_from = Sample, values_from = count) %>%
    column_to_rownames("Name") %>%
    as.matrix()

  ## Remove all-zero rows and columns
  count_mat <- count_mat[rowSums(count_mat) > 0, , drop = FALSE]
  count_mat <- count_mat[, colSums(count_mat) > 0, drop = FALSE]

  if (ncol(count_mat) < 2) {
    stop("Not enough non-zero samples for Acropora")
  }

  coldata <- data.frame(
    sample = colnames(count_mat),
    row.names = colnames(count_mat)
  )

  ## ---------------------------
  ## DESeq2 normalization
  ## ---------------------------
  dds <- DESeqDataSetFromMatrix(
    countData = round(count_mat),
    colData   = coldata,
    design    = ~ 1
  )

  dds <- estimateSizeFactors(dds, type = "poscounts")

  norm_counts <- counts(dds, normalized = TRUE)

  ## ---------------------------
  ## Filter low-expression miRNAs
  ## ---------------------------
  keep <- rowMeans(norm_counts) >= min_mean
  norm_counts <- norm_counts[keep, , drop = FALSE]

  ## ---------------------------
  ## Long format + CV by TP
  ## ---------------------------
  norm_counts %>%
    as.data.frame() %>%
    rownames_to_column("Name") %>%
    pivot_longer(
      cols = -Name,
      names_to = "Sample",
      values_to = "norm_count"
    ) %>%
    mutate(
      TP = str_extract(Sample, "TP[1-4]"),
      Species = "Acropora"
    ) %>%
    group_by(Name, TP) %>%
    summarise(
      mean_expr = mean(norm_count, na.rm = TRUE),
      sd_expr   = sd(norm_count, na.rm = TRUE),
      cv_expr   = sd_expr / mean_expr,
      n_samples = n(),
      .groups = "drop"
    )
}

cv_apul <- calc_cv_acropora(apul_mirna_results_biomin_long)
```

Look at CV, regardless of timepoint 
```{r cv-all-samples}
cv_summary <- cv_apul %>%
  summarise(
    n_mirna   = n_distinct(Name),
    mean_cv   = mean(cv_expr, na.rm = TRUE),
    median_cv = median(cv_expr, na.rm = TRUE),
    sd_cv     = sd(cv_expr, na.rm = TRUE),
    iqr_cv    = IQR(cv_expr, na.rm = TRUE),
    q25       = quantile(cv_expr, 0.25, na.rm = TRUE),
    q75       = quantile(cv_expr, 0.75, na.rm = TRUE)
  )

p1 <- ggplot(cv_apul, aes(y = cv_expr, x = "")) +
  geom_violin(alpha = 0.7, trim = TRUE, fill = "#E64B35") +
  geom_boxplot(
    width = 0.15,
    fill = "white",
    outlier.size = 0.5
  ) +
  labs(
    title = "miRNA Expression Variability (CV)",
    subtitle = "Acropora — higher CV indicates more variable expression",
    y = "Coefficient of Variation (CV)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(p1)
```

By timepoint
```{r}
ggplot(cv_apul, aes(x = TP, y = cv_expr)) +
  geom_violin(alpha = 0.7, trim = TRUE, fill = "#E64B35") +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
  labs(
    title = "miRNA Expression Variability by Time Point",
    subtitle = "Acropora",
    x = "Time Point",
    y = "Coefficient of Variation (CV)"
  ) +
  theme_minimal()

```

# Plot counts of miRNAs and their target biomin genes 

Filter biomin gene counts for the ones that are interacting with miRNAs and format these gene counts to long 
```{r pivot gene counts}
apul_mrna_biomin_long <- apul_biomin %>%
  select(-group_id) %>%
  filter(gene_id %in% apul_mir_interact_biomin$gene_id) %>%
  pivot_longer(
    -gene_id,
    names_to = "Sample",
    values_to = "count") %>%
  mutate(TP = str_extract(Sample, "TP[1-4]"))
```

Create interaction key 
```{r}
mir_mrna_map <- apul_mir_interact_biomin %>%
  select(miRNA, gene_id) %>%
  distinct()
```

Prep miRNA expression
```{r}
mir_expr <- apul_mirna_results_biomin_long %>%
  dplyr::rename(miRNA = Name) %>%
  mutate(entity = "miRNA")
```

Prep mRNA expression 
```{r}
mrna_expr <- apul_mrna_biomin_long %>%
  mutate(entity = "mRNA")
```

Join expression to interaction
```{r}
interaction_expr <- mir_mrna_map %>%
  left_join(mir_expr, by = "miRNA") %>%
  left_join(mrna_expr, by = "gene_id", suffix = c("_mir", "_mrna"))
```

Build df for plotting 
```{r}
plot_df <- bind_rows(
  # miRNA expression
  mir_expr %>%
    semi_join(mir_mrna_map, by = "miRNA") %>%
    transmute(
      regulator = miRNA,
      target = miRNA,
      TP,
      count,
      entity
    ),

  # mRNA expression
  mrna_expr %>%
    semi_join(mir_mrna_map, by = "gene_id") %>%
    left_join(mir_mrna_map, by = "gene_id") %>%
    transmute(
      regulator = miRNA,
      target = gene_id,
      TP,
      count,
      entity
    )
)
```

Summarize counts by TP
```{r}
summary_df <- plot_df %>%
  group_by(regulator, target, entity, TP) %>%
  summarise(
    mean_count = mean(count, na.rm = TRUE),
    sd_count   = sd(count, na.rm = TRUE),
    n          = n(),
    se_count   = sd_count / sqrt(n),
    .groups = "drop"
  )
```

Plot one miRNA
```{r}
plot_one_mirna_summary <- function(df, mirna_id) {

  ggplot(
    df %>% filter(regulator == mirna_id),
    aes(x = TP, y = mean_count, color = entity)
  ) +
    geom_point(
      position = position_dodge(width = 0.4),
      size = 3
    ) +
    geom_errorbar(
      aes(ymin = mean_count - se_count,
          ymax = mean_count + se_count),
      width = 0.2,
      position = position_dodge(width = 0.4)
    ) +
    #scale_y_log10() +
    labs(
      title = paste("miRNA–mRNA Expression Dynamics:", mirna_id),
      subtitle = "Mean ± SE across samples",
      x = "Time Point",
      y = "Raw Read Count",
      color = "Molecule"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
}
plot_one_mirna_summary(summary_df, "Cluster_1819")
```

Let's look at the conserved miRNAs (obtaining miRNA cluster names from this [file](https://github.com/urol-e5/timeseries-molecular-calcification/blob/main/D-Apul/output/04-Apul-sRNA-discovery-ShortStack/ShortStack_out/Results.txt) from shortstack output): 

- miR-100 (Cluster_17192)
- miR-2023 (Cluster_4752)
- miR-2025 (Cluster_14146)
- miR-2036 (Cluster_17245)

```{r}
plot_one_mirna_summary(summary_df, "Cluster_17192")
plot_one_mirna_summary(summary_df, "Cluster_4752")
plot_one_mirna_summary(summary_df, "Cluster_14146")
plot_one_mirna_summary(summary_df, "Cluster_17245")
```

Very interesting trends...the miRNAs tend to have greater expression than their target biomineralization genes across all timepoints. 

Write function so that one miRNA and its targets will be plotted per page and saved as a PDF
```{r}
plot_mirna_page <- function(df, mirna_id) {

  ggplot(
    df %>% filter(regulator == mirna_id),
    aes(x = TP, y = mean_count)) +
    # mRNA targets
    geom_point(
      data = ~ subset(.x, entity == "mRNA"),
      aes(color = "mRNA"),
      alpha = 0.6,
      size = 2) +
      #position = position_jitter(width = 0.05)
    geom_errorbar(
      data = ~ subset(.x, entity == "mRNA"),
      aes(
        ymin = mean_count - se_count,
        ymax = mean_count + se_count,
        color = "mRNA"),
      alpha = 0.5,
      width = 0.15) +

    # miRNA
    geom_point(
      data = ~ subset(.x, entity == "miRNA"),
      aes(color = "miRNA"),
      size = 2) +
    geom_errorbar(
      data = ~ subset(.x, entity == "miRNA"),
      aes(
        ymin = mean_count - se_count,
        ymax = mean_count + se_count,
        color = "miRNA"),
      alpha = 0.5,
      width = 0.15) +

    #scale_y_log10() +
    scale_color_manual(
      values = c("miRNA" = "#E64B35", "mRNA" = "#4DBBD5")) +
    labs(
      title = paste("miRNA–mRNA Expression Dynamics:", mirna_id),
      subtitle = "Mean ± SE across samples",
      x = "Time Point",
      y = "Raw Read Count",
      color = "Molecule") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "right")}

mirnas_to_plot <- sort(unique(summary_df$regulator))

# Save 
pdf(
  file = "../output/58-Hyp-miRNA-expression/Apul-mRNA-miRNA-biomin-interactions-across-TP.pdf",
  width = 8,
  height = 6)
for (mir in mirnas_to_plot) {
  p <- plot_mirna_page(summary_df, mir)
  print(p)}
dev.off()
```

Z-score counts for mRNA and miRNAs and plot again 
```{r}
summary_df_z <- summary_df %>%
  group_by(regulator, entity) %>%
  mutate(
    mean_count_z = as.numeric(scale(mean_count)),
    se_count_z   = se_count / sd(mean_count, na.rm = TRUE)
  ) %>%
  ungroup()

plot_mirna_page <- function(df, mirna_id) {

  ggplot(
    df %>% filter(regulator == mirna_id),
    aes(x = TP, y = mean_count_z)
  ) +

    # mRNA targets
    geom_point(
      data = ~ subset(.x, entity == "mRNA"),
      aes(color = "mRNA"),
      alpha = 0.6,
      size = 2
    ) +
    geom_errorbar(
      data = ~ subset(.x, entity == "mRNA"),
      aes(
        ymin = mean_count_z - se_count_z,
        ymax = mean_count_z + se_count_z,
        color = "mRNA"
      ),
      alpha = 0.5,
      width = 0.15
    ) +

    # miRNA
    geom_point(
      data = ~ subset(.x, entity == "miRNA"),
      aes(color = "miRNA"),
      size = 2
    ) +
    geom_errorbar(
      data = ~ subset(.x, entity == "miRNA"),
      aes(
        ymin = mean_count_z - se_count_z,
        ymax = mean_count_z + se_count_z,
        color = "miRNA"
      ),
      alpha = 0.5,
      width = 0.15
    ) +

    scale_color_manual(
      values = c("miRNA" = "#E64B35", "mRNA" = "#4DBBD5")
    ) +
    labs(
      title = paste("miRNA–mRNA Expression Dynamics:", mirna_id),
      subtitle = "Z-scored mean expression (± SE)",
      x = "Time Point",
      y = "Z-score",
      color = "Molecule"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "right"
    )
}

mirnas_to_plot_z <- sort(unique(summary_df_z$regulator))

# Save 
pdf(
  file = "../output/58-Hyp-miRNA-expression/Apul-mRNA-miRNA-biomin-interactions-across-TP-z-score.pdf",
  width = 8,
  height = 6)
for (mir in mirnas_to_plot_z) {
  p <- plot_mirna_page(summary_df_z, mir)
  print(p)}
dev.off()
```


Make network plot 
```{r}
# Set unique edges 
edges <- summary_df %>%
  filter(entity == "mRNA") %>%          # only miRNA → mRNA edges
  select(regulator, target) %>%
  distinct()

# Build node list 
nodes <- bind_rows(
  edges %>% transmute(name = regulator, type = "miRNA"),
  edges %>% transmute(name = target, type = "mRNA")) %>%
  distinct()

# Create igraph object 
g <- graph_from_data_frame(
  d = edges,
  vertices = nodes,
  directed = FALSE)

# Network
V(g)$color <- ifelse(V(g)$type == "miRNA", "#E64B35", "#4DBBD5")
V(g)$shape <- ifelse(V(g)$type == "miRNA", "square", "circle")
V(g)$size  <- ifelse(V(g)$type == "miRNA", 8, 5)

if ("weight" %in% edge_attr_names(g)) {
  E(g)$width <- scales::rescale(E(g)$weight, to = c(0.5, 3))
} else {
  E(g)$width <- 1
}

set.seed(1)
plot(
  g,
  layout = layout_with_fr(g),
  vertex.label.cex = 0.6,
  vertex.label.color = "black",
  main = "miRNA–mRNA Interaction Network (Acropora)"
)
```

